Gene-Environment Pathways to Cognitive Development and Psychotic-Like Experiences in Children

Version 1

Abstract

Abstract

Background

In children, psychotic-like experiences (PLEs) are related to risk of psychosis, schizophrenia, and other mental disorders. Maladaptive cognitive functioning is a well-known risk factor and early marker for psychosis, schizophrenia, and other mental disorders. Since cognitive functioning is linked to various genetic and environmental factors during development, we hypothesize that it mediates the effects of those factors on childhood PLEs. Using large, representative, longitudinal data, we tested the relationships of genetic and environmental factors (such as familial and neighborhood environment) with cognitive intelligence and their relationships with current and future PLEs in children.

Methods

To estimate unbiased associations against potential confounding variables, we leveraged large-scale, representative, multimodal data of 6,602 children (aged 9-10 years old; 47.15% females; 5,211 European-ancestry) from the Adolescent Brain and Cognitive Development Study. Linear mixed model and a novel structural equation modeling (SEM) method that allows unbiased estimation of both components and factors were used to estimate the joint effects of cognitive capacity polygenic scores (PGSs), familial and neighborhood socioeconomic status (SES), and supportive environment on NIH Toolbox cognitive intelligence and PLEs. We adjusted for ethnicity (genetically defined), schizophrenia PGS, and additionally unobserved confounders (using computational confound modeling).

Results

We identified that lower cognitive intelligence and higher PLEs correlated significantly with several genetic and environmental variables: i.e., lower PGSs for cognitive capacity, lower familial SES, lower neighborhood SES, lower supportive parenting behavior, and lower positive school environment. In SEM, lower cognitive intelligence significantly mediated the genetic and environmental influences on higher PLEs (Indirect effects of PGS: β range=-0.0355∼ -0.0274; Family SES: β range=-0.0429∼ -0.0331; Neighborhood SES: β range=0.0126∼ 0.0164; Positive Environment: β range=-0.0039∼ -0.003). Supportive parenting and a positive school environment had the largest total impact on PLEs (β range=-0.152∼ -0.1316) than genetic or environmental factors.

Conclusions

Our results reveal the role of genetic and environmental factors on children’s risk for psychosis via its negative impact on cognitive intelligence. Our findings have policy implications in that improving the school and family environment and promoting local economic development might be a way to enhance cognitive and mental health in children.

Key Points

While existing research shows the association between cognitive decline and the onset of psychosis, the genetic and environmental pathways to cognitive intelligence and psychotic risk in children remain unclear.

We identified the significant role of genetic and environmental factors (family, neighborhood, and school) on children’s risk for psychosis via a negative impact on cognitive intelligence.

Obtaining unbiased estimation by leveraging large, representative samples with multimodal data and advanced computational modeling for confounders, our results underscore the importance of incorporating socioeconomic policies into children’s cognitive and mental health programs.

Introduction

Childhood is the critical developmental period in human life. Cognitive development and mental health in this period significantly impact key life outcomes at later ages, including academic performance, economic productivity, physical health, intelligence, and psychopathology (Shonkoff, 2012; Walker et al., 2021). Literature shows the significant impact of social adversities on cognitive development and mental health in early childhood. Lower family socioeconomic status (SES), particularly household income, is linked to lower or delayed neurocognitive development and higher risk of psychopathology in childhood (Hair, Hanson, Wolfe, & Pollak, 2015; Noble et al., 2015; Peverill et al., 2021; Tomasi & Volkow, 2021; Weissman, Conger, Robins, Hastings, & Guyer, 2018).

Additional to family SES, the importance of neighborhood social environment on children’s neurocognitive development has been also highlighted (Gard et al., 2021; Tooley et al., 2020). Adverse neighborhood environment, such as the percent of families below poverty line, education level, and violence exposure, was associated with lower cognitive performance and greater risk for psychosis in children (Butler, Yang, Laube, Kühn, & Immordino-Yang, 2018; Karcher, Schiffman, & Barch, 2021; Rakesh, Seguin, Zalesky, Cropley, & Whittle, 2021; Taylor, Cooper, Jackson, & Barch, 2020). Conversely, as a protective factor against familial and neighborhood socioeconomic challenges, supportive parenting (Brody et al., 2017; Brody et al., 2019; Holmes et al., 2018; J. L. Luby et al., 2012; Joan L. Luby, Belden, Harms, Tillman, & Barch, 2016) and positive school environment (Gard et al., 2021; Piccolo, Merz, & Noble, 2019; Rakesh et al., 2021) have been highlighted to improve child cognition and mental health.

Psychotic-like experiences (PLEs) are prevalent in childhood and indicate the risk of psychosis and schizophrenia (van der Steen et al., 2019; Van Os & Reininghaus, 2016) and other psychopathologies including mood, anxiety, and substance disorders (Poulton et al., 2000; van der Steen et al., 2019). PLEs are linked to a lower cognitive intelligence (M. Cannon et al., 2002; Kelleher & Cannon, 2011). In childhood, PLEs also have a stronger association with environmental risk factors than any other internalizing/externalizing symptoms (Karcher, Schiffman, et al., 2021).

Maladaptive cognitive intelligence may precede the manifestation of psychotic symptoms (T. D. Cannon et al., 2000; Keefe et al., 2006; Reichenberg et al., 2005). Family studies show a decline in cognitive intelligence preceding psychotic symptoms is related to genetic risk (Cosway et al., 2000; Curtis, Calkins, Grove, Feil, & Iacono, 2001). In more recent studies these associations lead to the model that cognitive intelligence mediates the genetic risk for psychopathology and PLEs (Karcher, Paul, et al., 2021; Pat et al., 2021). Moreover, genetic overlap between psychotic disorders and cognitive capacity has been shown (Escott-Price et al., 2020; Le Hellard et al., 2016).

To obtain unbiased estimates of environmental effects, it is crucial to adjust for genetic confounding (Sariaslan et al., 2016) considering the large genetic effects on intelligence (Bouchard & McGue, 1981; Deary, Spinath, & Bates, 2006; Plomin & Spinath, 2004) and psychosis (Hilker et al., 2018; Kendler, 1983; Sullivan, Kendler, & Neale, 2003). Recent advances in genetics have led to the development of the polygenic score (PGS) approach: a computational method to estimate the genetic loading for a complex trait using statistical associations of each single-nucleotide polymorphism (SNP) identified by genome-wide association studies (GWAS) (Choi, Mak, & O’Reilly, 2020). In particular, PGS of cognitive capacity holds particular importance. Cognitive capacity PGS is positively correlated with cognitive intelligence and education attainment (Judd et al., 2020; Okbay et al., 2022; Plomin & von Stumm, 2018), brain morphometry, and negatively correlated with psychopathologies (e.g. psychosis, autism, depression, Alzheimer’s disease) and cognitive decline (Harden & Koellinger, 2020; Judd et al., 2020; Karcher, Paul, et al., 2021; Plomin & von Stumm, 2018). An important gap in the literature is the lack of integrated assessment of the effects of genetic and environmental factors at multiple levels (e.g., familial vs neighborhood) to dissect the genetic and environmental effects underlying abnormal cognitive intelligence and the PLEs. Addressing this with large and representative data will allow for a more complete understanding of the factors related to the development of psychosis and schizophrenia.

In this study, we firstly aim to assess the associations of cognitive PGSs, family and neighborhood SES, and supportive environment with children’s cognitive intelligence and PLEs using statistical and computational approaches to carefully consider potential confounding. We then aim to test whether intelligence has a mediating effect on the relationship between genetic and environmental factors and PLEs.

Materials and Methods

Study Participants

We used the multi-modal genetic and environmental data of 11,878 preadolescent children aged 9-10 years old collected from 21 research sites of the Adolescent Brain Cognitive Development (ABCD) Study, one of the largest longitudinal studies for children’s neurodevelopment in the United States. We analyzed the baseline, first year, and second year follow-up datasets included in ABCD Release 4.0, downloaded on January 25, 2022. After k-nearest neighbor imputation of missing values of covariates (4.67% of total observations), we removed participants with missing data on study variables. The final samples included 6,602 multiethnic children, which comprised 890 of African ancestry (13.48%), 91 of East Asian ancestry (1.38%), 5,211 of European ancestry (78.93%), 229 of Native American ancestry (3.47%), and 181 not specified (2.74%).

Data

NIH Toolbox Cognitive Performance

To assess children’s neurocognitive abilities, we used baseline observations of uncorrected composite scores of the total, fluid, and crystallized intelligence from the NIH Toolbox Cognitive Battery, which has seven cognitive instruments for examining executive function, episodic memory, language abilities, processing speed, working memory, and attention (Thompson et al., 2019).

Psychotic-Like Experiences (PLEs)

Baseline, first year, and second year follow-up of PLEs were measured using the children’s responses to the Prodromal Questionnaire-Brief Child Version. In line with previous research (Karcher et al., 2018; Karcher, Perino, & Barch, 2020; Karcher, Schiffman, et al., 2021), we computed Total Score and Distress Score, each indicating the number of psychotic symptoms and levels of total distress. Considering self-reports and parent-reports of psychopathology may differ (Achenbach, 2006), we additionally used parent-rated PLEs derived from four items of the Child Behavior Checklist according to previous studies (Karcher et al., 2018; Karcher et al., 2020; Karcher, Schiffman, et al., 2021).

Polygenic Scores (PGS)

To investigate the aggregated effect of genetic components, we estimated PGS of two representative cognitive capacity traits for each participant: education attainment (EA) and cognitive performance (CP) (Choi et al., 2020). We used the summary statistics released from the largest GWAS (Lee et al., 2018) of European-descent individuals for educational attainment (EA, n=1,131,881) and cognitive performance (CP, n=257,841). EA was measured as the years of schooling; CP was measured as the respondent’s score on cognitive ability assessments. To construct PGS of schizophrenia for sensitivity analyses, we used the summary statistics from the multiple GWAS of European sample (n=65,967; Ruderfer et al. (2018)) and East Asian sample (n=58,140; Lam et al. (2019)). We applied PRS-CSx, a high-dimensional Bayesian regression framework that improves cross-population prediction via continuous shrinkage prior to SNP effect sizes (Ge, Chen, Ni, Feng, & Smoller, 2019) (For details, see Appendix S1).

Family-, Neighborhood-, and School-level Environment

We assessed children’s family-level SES with family income, parental education, and family’s financial adversity based on parent self-reporting (Karcher et al., 2020; Taylor et al., 2020; Tomasi & Volkow, 2021). Higher family income and parental education and lower family’s financial adversity denote a higher family SES.

Neighborhood-level SES was assessed using the Area Deprivation Index (ADI), the percentage of individuals below -125% of the poverty level (henceforth “poverty”), and years of residence, which were associated with PLEs in prior research (Karcher, Schiffman, et al., 2021). Higher values of ADI and poverty and fewer years of residence indicate a lower neighborhood SES.

Based on existing literature (Karcher, Schiffman, et al., 2021; Rakesh et al., 2021), we measured the level of supportive parenting behavior (henceforth “good parenting”) and benign school environment (henceforth “good schooling”) to assess the effect of positive family and school environment on each individual.

Statistical Modeling

Linear Mixed Models

We used linear mixed models to minimize potential population stratification from different household environments and geographical locations (Choi et al., 2020). To avoid multicollinearity, CP and EA PGSs were used separately in two other models. Key variables of interest were PGS, family SES, neighborhood SES, and positive environment. As a random factor, the variable indicating ABCD research sites was used. Variables within each model had no signs of multicollinearity (Fox & Monette, 1992) (generalized variance inflation factor < 2.0 for every variable in all models). The child’s sex, age, genetic ethnicity, BMI, marital status of the caregiver, and family history of psychiatric disorders were included as covariates in the model (Karcher, Schiffman, et al., 2021). All continuous variables were standardized (z-scaled) beforehand to get standardized estimates, and the analyses were conducted with the lme4 package in R version 4.1.2. Throughout this paper, threshold for statistical significance was set at p<0.05, with correction for multiple comparisons using the false discovery rate. 95% bootstrapped confidence intervals were obtained with 5,000 iterations.

Path Modeling

To test the plausibility of whether cognitive intelligence may mediate the association between genetic and environmental factors and PLEs, we used an up-to-date SEM method, integrated generalized structured component analysis (IGSCA) (Hwang, Cho, Jung, et al., 2021). This approach is suited to our study using the multi-modal genetic and environmental variables in that it estimates models with both factors and components as statistical proxies for the constructs. We considered path-analytic relationships among the six key constructs: the PGS, family SES, neighborhood SES, positive family and school environment, general intelligence, and PLEs. As the PGSs, family SES, neighborhood SES, positive family and school environment, and PLEs may be seen as aggregations of their relevant observed variables, indicating that they are determined by their observed variables, we statistically represented these constructs as the weighted sums of their observed variables or components (Cho, Sarstedt, & Hwang, 2021). On the other hand, we represented general intelligence as a common factor that determines the underlying covariance pattern of fluid and crystallized intelligence, based on the classical g theory of intelligence (Jensen, 1998; Spearman, 1904).

The IGSCA model included the same covariates used in the linear mixed model as well as the ABCD research site as an additional covariate. We applied GSCA Pro 1.1 (Hwang, Cho, & Choo, 2021) to fit the IGSCA model to the data and checked the model’s goodness-of-fit index (GFI) (Jöreskog & Sörbom, 1986), standardized root mean square residual (SRMR), and total variance of all indicators and components explained (FIT) to assess its overall goodness-of-fit. A larger FIT value indicates more variance of all variables is explained by the specified model. The rules-of-thumb cutoff criteria in IGSCA is GFI≥0.93 and SRMR≤0.08 for an acceptable fit (Cho, Hwang, Sarstedt, & Ringle, 2020). Finally, we conducted conditional process analyses to investigate further the indirect and total effects of the constructs in the model. We obtained 95% confidence intervals based on 5,000 bootstrap samples to test the statistical significance of parameter estimates.

Sensitivity Analyses

To ensure robustness of the main analyses results, we conducted multiple sensitivity analyses. As the European-descent-based GWAS was used for constructing PGS, we reran the main analyses using participants of European ancestry (n=5,211) to adjust for ethnic confounding. Next, we tested effects of gene x environment interactions on cognitive intelligence and PLEs, respectively. We also assessed whether the effects of cognitive capacity PGS in the linear mixed model are significant after adjusting for schizophrenia PGS, a more direct genetic predictor of PLEs. Lastly, we adjusted for unobserved confounding bias in the linear mixed model, using a recently developed framework for causal inference based on null treatments assumption (Miao, Hu, Ogburn, & Zhou, 2022). As this method allows dependence among treatments, it can be utilized in examining the effects of multiple treatment variables having shared variances due to unobserved confounders (Okbay et al., 2022) (e.g., cognitive capacity PGS, family and neighborhood SES, positive family and school environment).

Results

Demographics

Table 1 presents the demographic characteristics of the final samples. For multiethnic subjects (main analyses, n=6,602), 47.15% were female, and the parents of 70.21% were married. In European ancestry samples (sensitivity analyses, n=5,211), 46.71% were female, and the parents of 77.47% were married. Children of European ancestry showed significantly different marital status (p<0.0001), lower BMI (p<0.0001) and family history of psychiatric disorders (p<0.0001) compared to children of other genetic ancestries.

Genetic influence on cognitive capacity correlates positively with intelligence and negatively with PLEs

As shown in Figure 1, higher PGSs of cognitive capacity phenotypes correlated significantly with higher intelligence (CP PGS: β=0.1113∼0.1793; EA PGS: β=0.0699∼0.1567). While CP PGS was associated only with lower baseline year Distress Score PLEs (β=-0.0323), EA PGS was associated with lower baseline year and follow-up PLEs of all measures (baseline: β=-0.0518∼-0.0519; 1-year: β=-0.0423∼-0.043; 2-year: β=- 0.036∼ -0.0463). No significant correlations were found between CP PGS and follow-up PLEs (p>0.05). The effects of EA PGS were larger on baseline year PLEs than follow-up PLEs. The effect sizes of EA PGS on children’s PLEs were twice as large as those of CP PGS (EA PGS: β=-0.036∼-0.0519; CP PGS: β=-0.0323). (Table S1).

Linear models testing genetic, socioeconomic, and environmental factors associated with cognitive intelligence and PLEs.

Standardized coefficients of a linear mixed model with CP PGS (A, B) and EA PGS (C, D). The analyses included 6,602 samples of multiethnicity. Cognitive intelligence and PLEs correlated with the PGSs, residential disadvantage, positive environment, and family SES in the opposite directions. Error bars indicate 95% bootstrapped confidence intervals with 5,000 iterations. CP and EA denote cognitive performance and education attainment, respectively; PGS, polygenic scores; SES, socioeconomic status; PLEs, psychotic-like experiences; ADI, Area Deprivation Index; Poverty, percentage of individuals below -125% of the poverty level; Years, years of residence. (*p-FDR<0.05)

Family and neighborhood SES correlates positively with intelligence and negatively with PLEs

Parental education associated positively with all types of intelligence (β=0.0699∼0.1745) and negatively with baseline year Total and Distress Score PLEs (β=- 0.0528∼-0.043), 1-year follow-up PLEs (β=-0.0538∼-0.0449), and 2-year follow-up PLEs (β=-0.0459∼-0.0389). Family income correlated positively with intelligence (β=0.0723∼0.1365) and negatively with 2-year follow-up parent-rated PLEs (β=-0.0503∼ - 0.0502). Family’s financial disadvantage correlated negatively with baseline year parent-rated PLEs (β=0.0726∼0.0728), 1-year follow-up PLEs of all types (β=0.0307∼0.0577), and 2-year follow-up PLEs of all types (β=0.0461∼0.0581).

The ADI correlated significantly negatively with all types of intelligence (β=- 0.0684∼-0.054). Additionally, a higher ADI correlated significantly with higher baseline year PLEs (β=0.0587∼0.0914), 1-year follow-up PLEs (β=0.0523∼0.0613), and 2-year follow-up PLEs (β=0.0397∼0.0449).

We found no significant associations of poverty with any of the target variables (p> 0.05). Years of residence correlated significantly with crystallized intelligence (β=0.035∼0.0372) and baseline year Total Score PLEs (β=-0.029∼-0.0273). (Table S1).

Positive family and school environments correlate positively with intelligence and negatively with the influence of PLEs

Good parenting behaviors showed significant negative correlations with baseline year PLEs (β=-0.0702∼-0.0419), 1-year follow-up PLEs (β=-0.0588∼-0.0397), and 2-year follow-up PLEs (β=-0.0623∼-0.0356) (Figure 1). Good school environment was associated positively with total intelligence (β=0.0353∼0.0397) and fluid intelligence (β=0.0514∼0.0545) and negatively with all three measures of baseline year PLEs (β=- 0.1193∼-0.0468), 1-year follow-up PLEs (β=-0.1078∼-0.04), and 2-year follow-up PLEs (β=- 0.1068∼-0.0586) (Table S1).

Structural Equation Modeling-Integrated Generalized Structured Component Analysis

The IGSCA model showed that intelligence mediated the effects of genes and environments on the risk for psychosis (PLEs) (Figure 2 and Table 2). The model showed a good model fit with a GFI of 0.9735, SRMR of 0.0359, and FIT value of 0.4912 (Cho et al., 2020). Intelligence was under significant direct influences of the cognitive capacity PGS (β=0.2427), family SES (β=0.2932), neighborhood SES (β=-0.1121), and positive environment (β=0.0268). Family SES and positive environment had significant negative direct effects on PLEs of all years. Cognitive capacity PGS and neighborhood SES showed no significant direct effects on any of the PLEs (p>0.05). Intelligence significantly mediated the effects of the PGS, family and neighborhood SES, and positive environment on PLEs of all years: the PGS (baseline year: β=-0.035; 1-year: β=-0.0355; 2-year: β=-0.0274), family SES (baseline year: β=-0.0423; 1-year: β=--0.0429; 2-year: β=-0.0331), neighborhood SES (baseline year: β=0.0162; 1-year: β=0.0164; 2-year: β=0.0126), and positive environment (baseline year: β=-0.0039; 1-year: β=-0.0039; 2-year: β=-0.003).

Pathway diagram of the structural equation model of gene-environment-intelligence-PLEs pathway: Cognitive capacity PGSs, family SES, neighborhood SES, positive environment, intelligence, and PLEs.

A. Standardized path coefficients are indicated on each path as direct effect estimates (significance level *p<0.05). B. Indirect pathways to PLEs via intelligence were significant for polygenic scores, family SES, neighborhood SES, and positive environment, indicating the significant mediating role of intelligence. C. Relative proportions of direct and indirect pathways within the total effects on PLEs. The standardized effect sizes of direct pathways are colored within each bar. In A-B, child sex, genetic ethnicity, BMI, marital status, family history of psychiatric disorders, and ABCD research sites were included as covariates. CP PGS and EA PGS denote polygenic scores of cognitive performance and education attainment, respectively; SES, socioeconomic status; PLEs, psychotic-like experiences; Crystallized and Fluid, crystallized and fluid intelligence; ADI, Area Deprivation Index; Poverty, percentage of individuals below -125% of the poverty level; Years, years of residence.

Note: * indicates a statistically significant parameter estimate at α=0.05.

For all observed years, positive environment had largest total effects on PLEs (baseline year: β=-0.152; 1-year: β=-0.1316; 2-year: β=-0.1364), followed by family SES (baseline year: β=-0.1216; 1-year: β=-0.1119; 2-year: β=-0.1164), neighborhood SES (baseline year: β=0.0504; 1-year: β=0.0329; 2-year: β=0.0192), and PGS (baseline year: β=- 0.0498; 1-year: β=-0.036; 2-year: β=-0.0365). The total effects of each indicator on PLEs were significant except for those of neighborhood SES (p>0.05).

Sensitivity Analyses

As sensitivity analyses, we reran our main analyses with adjustment for ethnic confounding, schizophrenia PGS, and unobserved confounding, respectively. Results of linear mixed model and IGSCA analyses were consistent with European ancestry participants (Table S3, S4). Interactions between PGS and good parenting had significant positive effects on 2-year follow-up Total and Distress Score PLEs (β=0.0343∼0.053) (Table S5). Unlike CP PGS and EA PGS, schizophrenia PGS did not have significant associations with any of the PLEs (p>0.05). Associations of cognitive capacitive PGS, family and neighborhood SES, and positive environment were similar after adjusting for schizophrenia PGS (See Table S6 for details). Lastly, when adjusting for unobserved confounders, CP and EA PGS, good parenting, and good schooling remained significantly correlated with intelligence and PLEs of all years (Table S7).

Discussion

This study investigated the relationships of the genetic and environmental influences on the development of intelligence and the PLEs in youth, leveraging genetic data from the large epidemiological samples and a multi-level environmental (socioeconomic) data. Our results support the model that genetic factors (PGS for cognitive capacity), socioeconomic conditions, and family and school environments may influence cognitive intelligence in children, and this impact may lead to the individual variability of the current and future PLEs in children. Our analysis with integrated data shows the contributions of genetic and environmental factors, respectively, to cognitive and mental wellness in children, and further provides policy implications to improve them.

Our SEM shows that cognitive intelligence mediates the environmental and genetic influence on the current and future PLEs. The environmental factors (family SES, neighborhood SES, and positive parenting and schooling) and PGS of cognitive capacity exhibit significant indirect effects via cognitive intelligence on PLEs. Prior studies identifying the mediation of cognitive intelligence focused on either genetic or environmental factors alone (Karcher, Paul, et al., 2021; Lewis et al., 2020). Our study advances this by demonstrating the specific and unbiased effects of genetic and environmental factors while controlling for each other. Considering the directions and magnitudes of the effects, though the effects of PGS remain significant, aggregated effects of environmental factors account for much greater degrees on PLEs (potential policy implications is discussed below). The indirect pathway of PGS & environment-intelligence-PLEs aligns with the studies with older clinical samples showing cognitive deficits as a precursor for the onset of psychosis and schizophrenia and their genetic and environmental factors (Eastvold, Heaton, & Cadenhead, 2007; Fett et al., 2020; Lewis et al., 2020; Vorstman et al., 2015).

Our results of cognitive intelligence mediating the genetic and environmental effects on PLEs may be related to several potential mechanisms. Children raised in higher family SES may have sufficient nutrition and cognitive stimulants, whereas children living in deprived neighborhoods may be exposed to higher rates of crime, air pollution, and substance abuse (Lewis et al., 2020; Marshall et al., 2020; Taylor et al., 2020; Tomasi & Volkow, 2021). Environmental enrichment may be associated with longer periods of neural plasticity (e.g. myelination, maturation of brain circuitry), leading to higher cognitive development and lower risk of mental disorders like psychosis (Tooley, Bassett, & Mackey, 2021). This may be further linked to the cognitive reserve theory. The theory suggests that genetic influence for cognitive capacity and environmental enrichment promotes more efficient, flexible brain networks, which may lead to greater resilience against psychopathology (Stern, 2009). Indeed, prior clinical studies show the linkage between cognitive reserve and psychosis (Amoretti et al., 2018; Leeson et al., 2011).

Our results indicate that genetic influences on cognitive capacity phenotypes are significantly linked to the risk of psychosis. While schizophrenia PGS showed no significant associations with PLEs, PGSs for CP and EA were strongly correlated with PLEs (baseline year, 1-year follow-up, and 2-year follow-up). These associations were robust after adjustment for ethnic confounding and unobserved confounders. Cognitive capacity PGS generally show higher predictive performance than PGS of any other traits (Lee et al., 2018; Okbay et al., 2022; Plomin & von Stumm, 2018). Genetic variants associated with CP and EA are related to complex traits across the life span, including neuroticism, depressive symptoms, smoking in adulthood, cognitive decline at a later age (Joo, Cha, Freese, & Hayes, 2022), risk for Alzheimer’s disease (Harden & Koellinger, 2020; Lee et al., 2018), brain volume, area, and thickness, as well as psychotic disorders (Karcher, Paul, et al., 2021). Prior gene expression studies suggest that polygenic signals for schizophrenia, bipolar disorder, and EA are significantly enriched in the central nervous system, particularly the cerebellum (Finucane et al., 2018). Our findings emphasize the importance of cognitive capacity PGS as a biomarker which not only implicates cognitive traits but also exhibits genetic overlap with the risk for psychosis.

The differing magnitudes of the PGS impact between EA and CP warrant attention. The effects of the EA PGS on the PLEs of all years were 160.68%∼371.67% larger than those of CP PGS. This discrepancy may result from that the larger sample size of EA GWAS than that of CP GWAS. Alternatively, the discrepancies in effect sizes may suggest different genetic compositions between EA and CP. Recent literature documents that more than half of the polygenic signal for EA is related to noncognitive and social skills required for successful educational attainment (Demange et al., 2021), whereas CP may rather be linked to cognitive skills. This observation also supports the well-established relationships of the EA PGS with socioeconomic and life-course outcomes (e.g., social mobility (Belsky et al., 2018), voter turnout (Aarøe et al., 2020), and wealth inequality (Barth, Papageorge, & Thom, 2020)), which may be influenced by unobserved environmental factors (Young, Benonisdottir, Przeworski, & Kong, 2019). In our analysis, using the two PGSs resulted in a more thorough assessment, leading to a more unbiased estimation of the genetic and environmental factors.

Furthermore, the significant effects of cognitive capacity PGS on cognitive intelligence (β=0.0699∼0.1793) remained robust and similar in magnitude after adjusting for ethnicity (β=0.0754∼0.1866) and other (unobserved) confounding (β=0.0546∼0.1776). As we controlled for family-, neighborhood-, and school-level environmental factors and unobserved confounders, our results may be interpreted as significant genetic influences on individual’s cognitive intelligence. This interpretation is supported by a recent study (Isungset et al., 2022): Despite of the socioeconomic differences in Norway (a typical social democratic welfare state) and the US (a typical liberal welfare state), the magnitudes of genetic influence on cognitive intelligence were similar (Norway: β=0.18; US: β=0.17). Cognitive capacity PGS is an important genetic factor across the nations and societies. Therefore, analyses omitting the genetic influence may be subject to overestimation of the socioeconomic impact (Plomin & von Stumm, 2018; Sariaslan et al., 2016).

This study shows that a high SES and positive environment, particularly good parenting and schooling, is associated with higher intelligence and a lower risk for PLEs in children. While prior research has emphasized the dominant role of family SES (e.g. family income) (Tomasi & Volkow, 2021), our SEM analyses(IGSCA) showed that positive environmental factors such as supportive parenting and schooling have a greater impact on children’s PLEs. Specifically, the effect sizes were the highest in supportive family and school environment, followed by family and neighborhood SES. Even after adjusting for ethnicity and unobserved confounders, the strong associations of good parenting and schooling with higher intelligence and fewer PLEs remained significant. These findings suggest that interventions that target positive family and school environments may be particularly effective. Recent research supports this notion, showing that interventions that promote supportive parenting and inclusive school environments can improve neurocognitive development, academic performance, and decrease risk behaviors such as drinking and emotional eating (Brody et al., 2017; Brody et al., 2019; Holmes et al., 2018).

Moreover, our results showed that positive parenting and schooling in baseline year were associated not only with baseline year PLEs but also with PLEs 1-2 years later. This is in line with prior research showing that intervention focused on parenting behavior and school environment have long-lasting positive effects that extend into adulthood and even across generations (Cunha & Heckman, 2007; Hill et al., 2020).

Our findings highlight the potential for preventative strategies that target residential environment, family SES, parenting, and schooling—a comprehensive approach that considers the entire ecosystem of children’s lives—to enhance children’s cognitive development and mental health. A holistic and quantitative approach that takes into account the comprehensive ecosystem of family, school, and residential environments may ensure treatment effects and efficient use of resources (Cree et al., 2018; Garner & Yogman, 2021; Shonkoff, 2012). For example, the Health Impact in 5 Years Initiative of the US Centers for Disease Control and Prevention (CDC, 2018) includes 14 evidence-based interventions, such as providing school-based prevention programs, public transportation, home improvement loans, and earned income tax credits, to tackle the social determinants of public health. Our study strengthens the idea that an interdisciplinary science-driven, coordinated approach to intervening in the select environmental factors may allow practical improvements in child development, particularly in those who are at a disadvantage.

Our study has some limitations. Firstly, due to data availability constraints in the ABCD study, we only utilized baseline observations for NIH Toolbox cognitive intelligence, and we could not test whether PLEs might be a mediator of intelligence. Secondly, the generalizability of our findings may be limited since most of the participants included in our analysis are from European ancestry. Finally, we did not include all important environmental variables, such as air pollution (Marshall et al., 2020) and social capital (Krabbendam, 2005) (not collected in the ABCD study).

In conclusion, our study provides unbiased estimation of the pathways of genetic factors of cognitive capacity and environmental factors of family, school, and neighborhood to cognitive and mental wellness in children. Our findings underscore the importance of a comprehensive approach that considers both biological and socioeconomic features in promoting young children’s cognitive development and mental health. Given the importance of child development, it requires joint efforts from multiple disciplines.

eLife Assessment

This study presents a useful inventory of the joint effects of genetic and environmental factors on psychotic-like experiences, and identifies cognitive ability as a potential underlying mediating pathway. The data were analyzed using solid and validated methodology based on a large, multi-center dataset. However, the claim that these findings are of relevance to psychosis risk and have implications for policy changes are only partially supported by the results.

Reviewer #1 (Public Review):

This study by Park et al. describes an interesting approach to disentangle gene-environment pathways to cognitive development and psychotic-like experiences in children. They have used data from the ABCD study and have included PGS of EA and cognition, environmental exposure data, cognitive performance data and self-reported PLEs. Although the study has several strengths, including its large sample size, interesting approach and comprehensive statistical model, I have several concerns:

- The authors have included follow-up data from the ABCD Study. However, it is not very clear from the beginning that longitudinal paths are being explored. It would be very helpful if the authors would make their (analysis) approach clearer from the introduction. Now, they describe many different things, which makes the paper more difficult to read. It would be of great help to see the proposed path model in a Figure and refer to that in the Method.

- There is quite a lot of causal language in the paper, particularly in the Discussion. My advice would be to tone this down.

- It's a bit unclear to me why the authors chose the PEs phenotype as their outcome of interest. They mainly speak about child development and mental health in general in the Introduction, so why focus on PLE? Aren't genes and environments also relevant for other types of mental health problems? Relatedly, in the Discussion the authors seem to conflate PLEs with psychosis. There is a large body of literature highlighting the differences between PLEs and psychosis, and this should - in my opinion - be adjusted throughout the introduction and discussion.

- I feel that the limitation section is a bit brief, and can be developed further.

- I like that the assessment of CP and self-reports PEs is of good quality. However, I was wondering which 4 items from the parent-reported CBCL were used and how did they correlate with the child-reported PEs? And how was distress taken into account in the child self-reported PEs measurement? Which PEs measures were used?

- What was the correlation between CP and EA PGSs?

- Regarding the PGS: why focus on cognitive performance and EA? It should be made clearer from the introduction that EA is not only measuring cognitive ability, but is also a (genetic) marker of social factors/inequalities. I'm guessing this is one of the reasons why the EA PGS was so much more strongly correlated with PEs than the CP PGS. See the work bij Abdellaoui and the work by Nivard.

- Considering previous work on this topic, including analyses in the ABCD Study, I'm not surprised that the correlation was not very high. Therefore, I don't think it makes a whole of sense to adjust for the schizophrenia PGS in the sensitivity analyses, in other words, it's not really 'a more direct genetic predictor of PLEs'.

- How did the FDR correction for multiple testing affect the results?

Overall, I feel that this paper has the potential to present some very interesting findings. However, at the moment the paper misses direction and a clear focus. It would be a great improvement if the readers would be guided through the steps and approach, as I think the authors have undertaken important work and conducted relevant analyses.

Reviewer #2 (Public Review):

This paper tried to assess the link between genetic and environmental factors on psychotic-like experiences, and the potential mediation through cognitive ability. This study was based on data from the ABCD cohort, including 6,602 children aged 9-10y. The authors report a mediating effect, suggesting that cognitive ability is a key mediating pathway in the link between several genetic and environmental (risk and protective) factors on psychotic-like experiences.

While these findings could be potentially significant, a range of methodological unclarities and ambiguities make it difficult to assess the strength of evidence provided.

Strengths of the methods:

The authors use a wide range of validated (genetic, self- and parent-reported, as well as cognitive) measures in a large dataset with a 2-year follow-up period. The statistical methods have the potential to address key limitations of previous research.

Weaknesses of the methods:

The rationale for the study is not completely clear. Cognitive ability is probably a more likely mediator of traits related to negative symptoms in schizophrenia, rather than positive symptoms (e.g., psychosis, psychotic-like symptom). The suggestion that cognitive ability might lead to psychotic-like symptoms in the general population needs further justification.

Terms are used inconsistently throughout (e.g., cognitive development, cognitive capacity, cognitive intelligence, intelligence, educational attainment...). It is overall not clear what construct exactly the authors investigated.

Not the largest or most recent GWASes were used to generate PGSes.

It is not fully clear how neighbourhood SES was coded (higher or lower values = risk?). The rationale, strengths, and assumptions of the applied methods are not fully clear. It is also not clear how/if variables were combined into latent factors or summed (weighted by what). It is not always clear when genetic and when self-reported ethnicity was used. Some statements might be overly optimistic (e.g., providing unbiased estimates, free even of unmeasured confounding; use of representative data).

It appears that citations and references are not always used correctly.

Strengths of the results:

The authors included a comprehensive array of analyses.

Weaknesses of the results:

Many results, which are presented in the supplemental materials, are not referenced in the main text and are so comprehensive that it can be difficult to match tables to results. Some of the methodological questions make it challenging to assess the strength of the evidence provided in the results.

Appraisal:

The authors suggest that their findings provide evidence for policy reforms (e.g., targeting residential environment, family SES, parenting, and schooling). While this is probably correct, a range of methodological unclarities and ambiguities make it difficult to assess whether the current study provides evidence for that claim.

Impact:

The immediate impact is limited given the short follow-up period (2y), possibly concerns for selection bias and attrition in the data, and some methodological concerns.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment:

This study presents a useful inventory of the joint effects of genetic and environmental factors on psychotic-like experiences, and identifies cognitive ability as a potential underlying mediating pathway. The data were analyzed using solid and validated methodology based on a large, multi-center dataset. However, the claim that these findings are of relevance to psychosis risk and have implications for policy changes are only partially supported by the results.

We appreciate the feedback and insightful suggestions from the editor and reviewers, which aided us to improve the manuscript. We believe the concerns initially raised were mostly due to areas that needed further clarification, which we have now clarified in this revised version. Our primary contribution lies in our meticulous analytical approach aimed at minimizing confounding effects and providing more precise estimates of the genetic and environmental impact on children's cognition and psychology. This method differs from the widely used general linear modeling in the field, which, in our opinion, may not be the optimal strategy for large-scale data analysis. Our comprehensive, tutorial-style description of the methods might serve as a valuable resource for the community.

Regarding the critique that our findings 'partially support the relevance to psychosis risk,' we have updated our manuscript to more accurately reflect this feedback. We have altered the narrative to indicate that psychotic-like experiences (PLE) are associated with the risk for psychosis, a connection substantiated by prior studies cited in our manuscript.

Similarly, in response to the comment that our findings 'partially support implications for policy changes,' we have nuanced our conclusion. However, we would like to emphasize our discovery that a negative genetic predisposition impacting cognitive development (i.e., low polygenic scores for cognitive phenotypes) can be counteracted by a positive school and familial environment. We believe that this finding could have meaningful implication for policy making and is robustly supported by our analyses.

We hope this revised manuscript more accurately reflects our research findings and its significances. Lastly, we would like to express our gratitude for your fair and detailed review process. Our experience working with eLife has been incredibly rewarding, and we commend your dedication to an encouraging and progressive publishing culture.

Public Reviews:

Reviewer #1

This study by Park et al. describes an interesting approach to disentangle gene-environment pathways to cognitive development and psychotic-like experiences in children. They have used data from the ABCD study and have included PGS of EA and cognition, environmental exposure data, cognitive performance data and self-reported PLEs. Although the study has several strengths, including its large sample size, interesting approach and comprehensive statistical model, I have several concerns:

The authors have included follow-up data from the ABCD Study. However, it is not very clear from the beginning that longitudinal paths are being explored. It would be very helpful if the authors would make their (analysis) approach clearer from the introduction. Now, they describe many different things, which makes the paper more difficult to read. It would be of great help to see the proposed path model in a Figure and refer to that in the Method.

We clarified the longitudinal paths tested in this study in Intro [line 149~159]. We also added a figure of the proposed path model (Figure 1) [Methods: line 231~238].

There is quite a lot of causal language in the paper, particularly in the Discussion. My advice would be to tone this down.

We adjusted and moderated the use of causal languages throughout the manuscript.

I feel that the limitation section is a bit brief, and can be developed further.

We clearly specified the limitations of our study. These included concerns about the representativeness of the ABCD samples, of the limited scope of longitudinal data, and the use of non-randomized, observational data [line 524~544].

I like that the assessment of CP and self-reports PEs is of good quality. However, I was wondering which 4 items from the parent-reported CBCL were used and how did they correlate with the child-reported PEs? And how was distress taken into account in the child self-reported PEs measurement? Which PEs measures were used?

Thanks for the clarification question. We report the Pearson’s correlation coefficients between the PLEs [line 198~200]. (The Reviewer #1 may have referred to the prior version of our manuscript submitted elsewhere, for this point has been already addressed in our initial submission to eLife).

What was the correlation between CP and EA PGSs?

The Pearson’s correlation between CP and EA PGS was 0.4331 (p<0.0001). We added the statistics to the manuscript. [line 214]

Regarding the PGS: why focus on cognitive performance and EA? It should be made clearer from the introduction that EA is not only measuring cognitive ability, but is also a (genetic) marker of social factors/inequalities. I'm guessing this is one of the reasons why the EA PGS was so much more strongly correlated with PEs than the CP PGS. See the work bij Abdellaoui and the work by Nivard.

We appreciate the reviewer’s insightful feedback. Acknowledging the role of both CP and EA PGSs in our study, we agree with the observation that EA PGS goes beyond gauging cognitive aptitude—it also serves as an indicator of societal influences and inequalities. The multifaceted nature of EA PGS could be the reason underlying the stronger correlation with PLEs compared to CP PGS. In response to this feedback, we revised our introduction to articulate the multifaceted role of EA PGS in more precise terms. For supporting our assertions, we have included references to prior studies (Abdellaoui et al., 2022) [line 131~142].

Abdellaoui, A., Dolan, C. V., Verweij, K. J. H., & Nivard, M. G. (2022). Gene–environment correlations across geographic regions affect genome-wide association studies. Nature Genetics. doi:10.1038/s41588-022-01158-0

Considering previous work on this topic, including analyses in the ABCD Study, I'm not surprised that the correlation was not very high. Therefore, I don't think it makes a whole of sense to adjust for the schizophrenia PGS in the sensitivity analyses, in other words, it's not really 'a more direct genetic predictor of PLEs'.

We thank the reviewer for the thoughtful comments. We acknowledge that the correlation between schizophrenia PGS and PLE may not be exceedingly high, as evidenced by previous work, including analyses from the ABCD study. However, we would like to emphasize our rationale for adjusting schizophrenia PGS in the sensitivity analyses. Our study design stemmed from the established associations between PLEs and increased risk for schizophrenia. Existing studies have reported significant associations between schizophrenia PGS and cognitive deficits in both psychosis patients (Shafee et al., 2018) and people at risk for psychosis (He et al., 2021). Notable, the PGS for schizophrenia has shown significant associations with PLEs, arguably more so than PGS for PLEs itself (Karcher et al., 2018). Our updated manuscript has incorporated these references to improve clarity. [line 307~309]. By adding this layer of adjustment, we believe that our mixed linear model more precisely examines the relationship between the cognitive phenotype PGS and PLEs, in terms of both sensitivity and specificity.

He, Q., Jantac Mam-Lam-Fook, C., Chaignaud, J., Danset-Alexandre, C., Iftimovici, A., Gradels Hauguel, J., . . . Chaumette, B. (2021). Influence of polygenic risk scores for schizophrenia and resilience on the cognition of individuals at-risk for psychosis. Translational Psychiatry, 11(1). doi:10.1038/s41398-021-01624-z

Karcher, N. R., Paul, S. E., Johnson, E. C., Hatoum, A. S., Baranger, D. A. A., Agrawal, A., . . . Bogdan, R. (2021). Psychotic-like Experiences and Polygenic Liability in the Adolescent Brain Cognitive Development Study. Biological Psychiatry: Cognitive Neuroscience and Neuroimaging. doi:https://doi.org/10.1016/j.bpsc.2021.06.012

Shafee, R., Nanda, P., Padmanabhan, J. L., Tandon, N., Alliey-Rodriguez, N., Kalapurakkel, S., . . . Robinson, E. B. (2018). Polygenic risk for schizophrenia and measured domains of cognition in individuals with psychosis and controls. Translational Psychiatry, 8(1). doi:10.1038/s41398-018-0124-8

How did the FDR correction for multiple testing affect the results?

Please note that we have clarified our FDR correction in the methods

As detailed in the method section [line 254~255], we applied False Discovery Rate (FDR) correction for multiple testing across nine key variables in the study: PGS (CP or EA), family income, parental education, family’s financial adversity, Area Deprivation Index, years of residence, proportion of population below -125% of the poverty line, positive parenting behavior, and positive school environment. An exception was made in our additional sensitivity analysis, where we included schizophrenia PGS in the linear mixed model for adjustment, thus the FDR correction was applied across ten key variables instead. Overall, the application of FDR correction had minimal impact on our findings. Most associations between the key variables and the outcomes that were originally marked as highly significant sustained their significance after the FDR correction.

Overall, I feel that this paper has the potential to present some very interesting findings. However, at the moment the paper misses direction and a clear focus. It would be a great improvement if the readers would be guided through the steps and approach, as I think the authors have undertaken important work and conducted relevant analyses.

We express our appreciation to the reviewer for the positive feedback and constructive suggestions, which only serve to improve and strengthen our manuscript. We have incorporated the suggested corrections and clarifications in response to the reviewer's suggestions. We believe that these changes will not only enhance the overall readability but also more effectively emphasize the significance and implication of our work.

Reviewer #2 (Public Review):

This paper tried to assess the link between genetic and environmental factors on psychotic-like experiences, and the potential mediation through cognitive ability. This study was based on data from the ABCD cohort, including 6,602 children aged 9-10y. The authors report a mediating effect, suggesting that cognitive ability is a key mediating pathway in the link between several genetic and environmental (risk and protective) factors on psychotic-like experiences.

While these findings could be potentially significant, a range of methodological unclarities and ambiguities make it difficult to assess the strength of evidence provided.

Strengths of the methods:

The authors use a wide range of validated (genetic, self- and parent-reported, as well as cognitive) measures in a large dataset with a 2-year follow-up period. The statistical methods have the potential to address key limitations of previous research.

Weaknesses of the methods:

The rationale for the study is not completely clear. Cognitive ability is probably a more likely mediator of traits related to negative symptoms in schizophrenia, rather than positive symptoms (e.g., psychosis, psychotic-like symptom). The suggestion that cognitive ability might lead to psychotic-like symptoms in the general population needs further justification.

We appreciate the reviewer’s concern regarding the role of cognitive ability in relation to schizophrenia symptoms. We are aware that cognitive ability often serves as a mediator of psychotic-like experiences. However, to our best knowledge, a growing body of research has proposed that cognitive ability can mediate positive symptoms in schizophrenia including psychotic-like experiences. The studies by Howes & Murray (2014) and Garety et al. (2001) suggested that deficits in cognitive ability can potentially contribute to the manifestation of positive symptoms such as psychotic-like experiences. We have elaborated on this aspect in the Introduction section [line 104-115].

Howes, O. D., & Murray, R. M. (2014). Schizophrenia: an integrated sociodevelopmental-cognitive model. The Lancet, 383(9929), 1677-1687. doi:https://doi.org/10.1016/S0140-6736(13)62036-X

Garety, P. A., Kuipers, E., Fowler, D., Freeman, D., & Bebbington, P. E. (2001). A cognitive model of the positive symptoms of psychosis. Psychological Medicine, 31(2), 189-195. doi:10.1017/S0033291701003312

Terms are used inconsistently throughout (e.g., cognitive development, cognitive capacity, cognitive intelligence, intelligence, educational attainment...). It is overall not clear what construct exactly the authors investigated.

We thank the reviewer’s feedback regarding the consistency of terminology in our manuscript. Per the suggestion, we standardized the use of ‘cognitive capacity’ and now consistently refer to it as ‘cognitive phenotypes’ throughout our manuscript. Furthermore, we explicitly stated in the Introduction section that our two PGSs of focus will be termed ‘cognitive phenotypes PGSs’, aligning with terminology used in prior studies (Joo et al., 2022; Okbay et al., 2022; Selzam et al., 2019) [line 140~142].

Joo, Y. Y., Cha, J., Freese, J., & Hayes, M. G. (2022). Cognitive Capacity Genome-Wide Polygenic Scores Identify Individuals with Slower Cognitive Decline in Aging. Genes, 13(8), 1320. doi:10.3390/genes13081320

Okbay, A., Wu, Y., Wang, N., Jayashankar, H., Bennett, M., Nehzati, S. M., . . . Young, A. I. (2022). Polygenic prediction of educational attainment within and between families from genome-wide association analyses in 3 million individuals. Nature Genetics, 54(4), 437-449. doi:10.1038/s41588-022-01016-z

Selzam, S., Ritchie, S. J., Pingault, J.-B., Reynolds, C. A., O’Reilly, P. F., & Plomin, R. (2019). Comparing Within- and Between-Family Polygenic Score Prediction. The American Journal of Human Genetics, 105(2), 351-363. doi:https://doi.org/10.1016/j.ajhg.2019.06.006

Not the largest or most recent GWASes were used to generate PGSes.

We appreciate the reviewer’s observation. Indeed, we were unable to utilize the most recent or the largest GWAS for cognitive performance, educational attainment, and schizophrenia due to the timeline of our study. Regrettably, the commencement of our study preceded the publication of the ‘currently’ the largest or most recent GWAS studies by Okbay et al. (2022) and Trubetskoy et al. (2022). Our research was conducted with the best available data at that time, which was the GWAS of European-descent individuals for educational attainment and cognitive performance (Lee et al, 2018). To eliminate any potential confusion, we adjusted the text to specify that our study used 'a GWAS of European-descent individuals for educational attainment and cognitive performance' rather than the largest GWAS [line 206~208].

It is not fully clear how neighbourhood SES was coded (higher or lower values = risk?). The rationale, strengths, and assumptions of the applied methods are not fully clear. It is also not clear how/if variables were combined into latent factors or summed (weighted by what). It is not always clear when genetic and when self-reported ethnicity was used. Some statements might be overly optimistic (e.g., providing unbiased estimates, free even of unmeasured confounding; use of representative data).

Thank you for pointing this out. Consistent with the illustration of neighborhood SES in the Methods, higher values of neighborhood SES indicate risk [line 217~228]. In the original Figure 2, higher value of neighborhood SES links to lower intelligence (direct effects: β=-0.1121) and higher PLEs (indirect effects: β=-0.0126~ -0.0162). We think such confusion might have been caused by the difference between family SES (higher values = lower risk) neighborhood SES (higher values = higher risk). Thus, we changed the terms to ‘High Family SES’ and ‘Low Neighborhood SES’ in the corrected figure (Figure 3) for clarification.

Considering that shorter duration of residence may be associated with instability of residency, it may indicate neighborhood adversity (i.e., higher risk). This definition of the ‘years of residence’ variable is in line with the previous study by Karcher et al. (2021).

During estimation, the IGSCA determines weights of each observed variable in such a way as to maximize the variances of all endogenous indicators and components. We added this explanation in the description about the IGSCA method [line 266~268].

We deleted overly optimistic statements like ‘unbiased estimates’ and used expressions such as ‘adjustment for observed/unobserved confounding’ instead, throughout our manuscript.

Karcher, N. R., Schiffman, J., & Barch, D. M. (2021). Environmental Risk Factors and Psychotic-like Experiences in Children Aged 9–10. Journal of the American Academy of Child & Adolescent Psychiatry, 60(4), 490-500. doi:10.1016/j.jaac.2020.07.003

It appears that citations and references are not always used correctly.

We thoroughly checked all citations and specified the references for each statement: We deleted Plomin & von Stumm (2018) and Harden & Koellinger (2020) and cited relevant primary studies (e.g., Lee et al., 2018; Okbay et al., 2022; Abdellaoui et al., 2022) instead. We also specified the references supporting the statement that educational attainment PGS links to brain morphometry (Judd et al., 2020; Karcher et al., 2021). As Okbay et al. (2022) use PGS of cognitive intelligence (which mentions the analyses results in their supplementary materials) as well as educational attainment, we decided to continue citing this reference [line 131~141].

Strengths of the results:

The authors included a comprehensive array of analyses.

We thank the reviewer for the positive comment.

Weaknesses of the results:

Many results, which are presented in the supplemental materials, are not referenced in the main text and are so comprehensive that it can be difficult to match tables to results. Some of the methodological questions make it challenging to assess the strength of the evidence provided in the results.

As you rightly identified, we inadvertently failed to reference Table S2 in the main text. We have since corrected this omission in the Results section for the IGSCA (SEM) analysis [line 376]. The remainder of the supplementary tables (Table S1, S3~S7) have been appropriately cited in the main manuscript. We recognize that the quantity of tables provided in the supplementary materials is substantial. However, given the comprehensiveness and complexity of our analyses, which encompass a wide array of study variables, these tables offer intricate results from each analysis. We deem these results, which include valuable findings from sensitivity analyses and confound testing, too significant to exclude from the supplementary materials. That said, we are open to, and would greatly welcome, any further suggestions on how to present our supplementary results in a more clear and digestible format. Your guidance in this matter is highly valued.

Appraisal:

The authors suggest that their findings provide evidence for policy reforms (e.g., targeting residential environment, family SES, parenting, and schooling). While this is probably correct, a range of methodological unclarities and ambiguities make it difficult to assess whether the current study provides evidence for that claim.

We believe that with the improvement we made in this revised manuscript, this concern may have been successfully mitigated.

Impact:

The immediate impact is limited given the short follow-up period (2y), possibly concerns for selection bias and attrition in the data, and some methodological concerns.

We appreciate the feedback provided in the reviewer's impact statement. We added as study limitations [line 524~544] that the impact of our findings may be limited due to the relatively short follow-up period, the possibility of sample selection bias, and the problems of interpreting results from an observational study as causality (despite the novel causal inference methods, designed for non-randomized, observational data, that we used).

As responded above (and also in more detail in the Reviewer #2’s Recommendations For The Authors section below), we made necessary corrections and clarifications for the points suggested by the reviewer. As we are willing to make additional revisions, please feel free to give comments if you feel that our corrections are insufficient or inappropriate.

Nevertheless, we would like to discuss some points. We sincerely hope this following response does not come across as argumentative to the reviewer and the editor. We fully understand the reviewer's perspective on this matter, and we agree that the issues raised about the ABCD study are absolutely valid. However, when evaluating the overall impact of a study, other factors, such as how the field has been assessing the impact of similar studies, should also be considered.

Firstly, the potential selection bias and attrition in the ABCD data may not necessarily limit the conclusions of this study. While recognizing the potential issues with the ABCD data is important, we feel that judging the impact of our findings as "limited" based on these issues may not be entirely fair. This is because no study, particularly those of a nationwide scale such as the UK Biobank, IMAGEN, HEAL, HBCD, etc., is completely free of limitations. Typically, the potential limitations of the data don't undermine the impact of individual studies' findings. Numerous studies using ABCD data have been published in top-tier journals—despite the limitations of the ABCD study—underscoring the scientific merit of the findings. For example, the study by Tomasi, D., & Volkow, N. D. (2021), entitled "Associations of family income with cognition and brain structure in USA children: prevention implications," published in Molecular Psychiatry, might be highly relevant to the limitations of the ABCD study raised by the reviewer. The scientific community, including editors, reviewers, and readers, may have appreciated the impact of this study despite the acknowledged limitations of the ABCD data.

Secondly, the two-year time window of our longitudinal analysis might not impact the aim of this study—an iterative assessment of the associations between genetic and environmental variables with cognitive intelligence and mental health, with a focus on PLE, in preadolescents. Had we aimed to test the developmental trajectory from childhood to adolescence, perhaps a longer timeframe would have made more sense. So, we do not agree with the reviewer’s assessment that the short time window limits the impact of our study.

Suggested revisions based on the combined reviewer feedback:

1. The terminology used should be carefully reviewed and revised

• Please use the correct terminology for the key concepts assessed in this study. For example, authors sometimes conflate PLEs and psychosis, two related but separate constructs. Furthermore, the terms 'good parenting' and 'good schooling' are vague and subjective.

• The authors use multiple terms to refer to cognitive ability (cognitive capacity, intelligence, cognitive intelligence, etc). The term 'cognitive development' in the title and manuscript does not seem to be justified given the focus on different measures of cognitive ability at a single time point (i.e. baseline).

• Please avoid causal language and using statements that cannot be entirely substantiated (e.g. unbiased estimates, free from unmeasured confounding)

Thank you for suggesting this point. We revised all key terminologies used throughout our manuscript.

Per your suggestion, we specified that PLEs indicate the risk of psychosis and often precede schizophrenia. We checked all misused cases of the term ‘psychosis’ and corrected them as ‘PLEs’. We also changed the terms 'good parenting' and 'good schooling' to ‘positive parenting behavior’ and ‘positive school environment’.

We changed the term ‘cognitive development’ to ‘cognitive ability’ throughout our manuscript. We also changed the title to ‘Gene-Environment Pathways to Cognitive Intelligence and Psychotic-Like Experiences in Children’ because we used ‘cognitive intelligence’ for NIH toolbox variable in the text.

We corrected and tone-downed all causal languages used in our manuscript. As mentioned by the reviewers, we deleted statements like ‘unbiased estimates’ and used expressions such as ‘adjustment for observed/unobserved confounding’ instead.

1. A stronger rationale for the focus on PLEs, and the potential mediating role of cognitive ability in genetic and environmental effects on PLES, should be provided

We appreciate the raised concerns that cognitive ability may serve as a mediator of psychotic-like experiences. To our best knowledge, it has been proposed that cognitive ability can be a mediator of positive symptoms in schizophrenia (including psychotic-like experiences), as well as negative symptoms. This mediating role of cognitive ability was proposed in several prior studies on cognitive model of schizophrenia/psychosis. Per your suggestion, we included an additional justification in Intro [line 104~115] where we highlighted that cognitive ability has been proposed as a potential mediator of genetic and environmental influence on positive symptoms of schizophrenia such as psychotic-like experiences. We refer to studies conducted by Howes & Murray (2014) and Garety et al. (2001).

Howes, O. D., & Murray, R. M. (2014). Schizophrenia: an integrated sociodevelopmental-cognitive model. The Lancet, 383(9929), 1677-1687. doi:https://doi.org/10.1016/S0140-6736(13)62036-X

Garety, P. A., Kuipers, E., Fowler, D., Freeman, D., & Bebbington, P. E. (2001). A cognitive model of the positive symptoms of psychosis. Psychological Medicine, 31(2), 189-195. doi:10.1017/S0033291701003312

1. As described in more detail by the reviewers, more information should be provided about the measures used in the study and how they relate to one another (e.g. correlations between PQ-BC and CBCL; PGS-CA and PGS-EA).

Thank you for your suggestion. Although this information was already provided in our initial submission, it appears that the Reviewer #1’s might have referred to the prior version of our manuscript submitted elsewhere before eLife.

To clarify, our findings reveal significant Pearson’s correlation coefficients between PLEs across all time-points (baseline year: r=0.095~0.0989, p<0.0001; 1-year follow-up: r=0.1322~0.1327, p<0.0001; 2-year follow-up: r= 0.1569~0.1632, p<0.0001) and we added this information in the Method section [line 198~200]. We also added the Pearson’s correlation between the two PGSs (r=0.4331, p<0.0001) in the Methods for PGS [line 214].

1. More details are needed regarding the analytical strategies used (e.g. how imputation was performed, why PGS were not based on the largest and most recent GWASes, whether latent or observed variables were examined, what exactly the supplementary materials show and how they relate to information provided in the main text).

We appreciate your feedback. We acknowledge the concerns about the GWAS sources utilized for the study. Unfortunately, our study commenced prior to the publication of the ‘currently’ most recent or largest GWAS by Okbay et al. (2022) and Trubetskoy et al. (2022). Our research was conducted with the best available data at that time, which was the largest GWAS of European-descent individuals for educational attainment and cognitive performance (Lee et al, 2018). We have now clarified this point in the manuscript. [line 206~208]

Also, we specified the use of composite indicators for the PGS, family SES, neighborhood SES, positive family and school environment, and PLEs, while latent factors were used for cognitive intelligence [line 269~285].

We highly appreciate the reviewer’s comments regarding the supplementary materials. We regret overlooking the citation of Table S2 in the main manuscript, and this has now been rectified in the Results section for the IGSCA (SEM) analysis [line 376]. The remaining supplementary tables (Table S1, S3~S7) have been correctly referenced within the manuscript. We acknowledge that the supplementary materials are extensive due to the comprehensive array of study variables and intricate results from each analysis. However, given that our analyses encompass a wide array of study variables, these tables offer intricate results from each analysis. We deem these results, which include valuable findings from sensitivity analyses and confound testing, too crucial to exclude from the supplementary materials. That said, we are open to any further suggestions to make our supplementary results more accessible and digestible. In order to improve the accessibility and clarity of our presentation, we are fully committed to making any necessary changes and look forward to any further recommendations.

1. The limitation section should be expanded and statements regarding the implications of the study findings should be qualified accordingly (e.g. short follow-up period, potential for attrition and selection bias, reverse causation, etc)

We specified additional potential constraints of our study, including limited representativeness, limited periods of follow-up data (baseline year, 1-year, and 2-year follow-up), possible sample selection bias, and the use of non-randomized, observational data [line 524~544].

1. Please ensure that the references provided support the statements in the text to which they are linked to.

Thank you for pointing this out. We thoroughly went over all citations and corrected the inaccurately or vaguely cited references for each statement.

Reviewer #2 (Recommendations For The Authors):

1. Please use terms consistently and correctly. E.g., 'cognitive capacity' is not the same as 'educational attainment'.

We thank the reviewer’s feedback regarding the consistency of terminology in our manuscript. Per the suggestion, we standardized the use of ‘cognitive capacity’ and now consistently refer to it as ‘cognitive phenotypes’ throughout our manuscript. Furthermore, we explicitly stated in the Introduction section that our two PGSs of focus will be termed ‘cognitive phenotypes PGSs’, aligning with terminology used in prior studies (Joo et al., 2022; Okbay et al., 2022; Selzam et al., 2019) [line 140~142].

Joo, Y. Y., Cha, J., Freese, J., & Hayes, M. G. (2022). Cognitive Capacity Genome-Wide Polygenic Scores Identify Individuals with Slower Cognitive Decline in Aging. Genes, 13(8), 1320. doi:10.3390/genes13081320

Okbay, A., Wu, Y., Wang, N., Jayashankar, H., Bennett, M., Nehzati, S. M., . . . Young, A. I. (2022). Polygenic prediction of educational attainment within and between families from genome-wide association analyses in 3 million individuals. Nature Genetics, 54(4), 437-449. doi:10.1038/s41588-022-01016-z

Selzam, S., Ritchie, S. J., Pingault, J.-B., Reynolds, C. A., O’Reilly, P. F., & Plomin, R. (2019). Comparing Within- and Between-Family Polygenic Score Prediction. The American Journal of Human Genetics, 105(2), 351-363. doi:https://doi.org/10.1016/j.ajhg.2019.06.006

1. The authors study 'cognitive performance using seven instruments', but it is not clear how fluid and crystalline intelligence was defined/operationalized.

Thank you for pointing this out. We specified the NIH Toolbox tests used for composite scores of fluid and crystallized intelligence, respectively. “We utilized baseline observations of uncorrected composite scores of fluid intelligence (Dimensional Change Card Sort Task, Flanker Test, Picture Sequence Memory Test, List Sorting Working Memory Test), crystallized intelligence (Picture Vocabulary Task and Oral Reading Recognition Test), and total intelligence (all seven instruments) provided in the ABCD Study dataset” [line 180~187].

1. I don't think Lee 2018 is the largest GWAS for educational attainment. That would be Okbay 2022. It needs to be described how cognitive performance was defined in Lee 2018. Why did the authors not use the Trubetskoy 2022 schizophrenia GWAS?

Thank you for mentioning this point. The reason why we were not able to use the largest GWAS for CP, EA and schizophrenia is because (unfortunately) our study started earlier than the point when the GWAS studies by Okbay et al. (2022) and Trubetskoy et al. (2022) were published. We corrected that our study used ‘a GWAS of European-descent individuals for educational attainment and cognitive performance’ instead of the largest GWAS [line 206~208].

1. It is unclear how neighbourhood SES was coded. The authors seem to suggest that higher values indicate risk, but Figure 2 suggests that higher values links to higher intelligence and lower PLE.

Thank you very much for pointing this out. Consistent with the illustration of neighborhood SES in the Methods section, higher values of neighborhood SES indicate risk. In the original Figure 2, higher values of neighborhood SES links to lower intelligence (direct effects: β=-0.1121) and higher PLEs (indirect effects: β=-0.0126~-0.0162). We think such confusion might have been caused by the difference between family SES (higher values = lower risk) neighborhood SES (higher values = higher risk). Thus, we changed the terms to ‘High Family SES’ and ‘Low Neighborhood SES’ in the corrected figure (Figure 3) for clarification.

1. Also, the 'year of residence' variable is unclearly defined. Does this mean that a shorter duration of residency (even in a good neighbourhood) indicate risk?

Thank you for mentioning this point. Considering that shorter duration of residence may be associated with instability of residency, it may indicate neighborhood adversity (i.e., higher risk). This definition of the ‘years of residence’ variable is in line with the previous study by Karcher et al. (2021).

Karcher, N. R., Schiffman, J., & Barch, D. M. (2021). Environmental Risk Factors and Psychotic-like Experiences in Children Aged 9–10. Journal of the American Academy of Child & Adolescent Psychiatry, 60(4), 490-500. doi:10.1016/j.jaac.2020.07.003

1. Please provide information on how correlated the two PGSes were.

Thank you for your suggestion. We added the Pearson’s correlation between the two PGSs (r=0.4331, p<0.0001) in the Methods section for PGS [line 214].

1. Information on the outcome variable in the 'linear mixed models' section is missing. I assumed it was PLE.

Thank you for notifying us of this point. We added the information on the outcome variables in the section for linear mixed models [line 242~244].

1. In the 'Path Modeling' section, please explain what 'factors and components' concretely refer to. How is this different from a standard SEM with latent factors?

Thank you for your comment on the need to elaborate the IGSCA method. We added that different from standard SEM methods which only uses latent factors, the IGSCA method can use components as well as latent factors as constructs in model estimation. This allows the IGSCA method to control bias more effectively in estimation compared to the standard SEM [line 261~268].

1. The sentence starting line 229 is unclear. Does this mean variables were not used to generate latent factors. And if not, what weights were used to create a 'weighted sum'?

Thank you for mentioning this point. The sentence means that we treated PGSs, family SES, neighborhood SES, positive family and school environment, and PLEs as composite indicators (derived from a weighted sum of relevant observed variables), while general intelligence was represented as a latent factor.

It has been suggested from prior studies that these variables (PGSs, family SES, neighborhood SES, positive family and school environment, and PLEs) are less likely to share a common factor and were assessed as a composite index during analyses. For instance, Judd et al. (2020) and Martin et al. (2015) analyze genetic influence of educational attainment and ADHD as composite indicators. Also, as mentioned in Judd et al. (2020), socioenvironmental influences are often analyzed as composite indicators. Studies on psychosis continuum (e.g., van Os et al., 2009) suggest that psychotic disorders are likely to have multiple background factors instead of having a common factor, and notes that numerous prior research uses composite indices to measure psychotic symptoms. Based on this literature, we used components for these variables.

The IGSCA determines weights of each observed variable to maximize the variances of the endogenous indicators and components [added in line 265~268].

On the other hand, we treated general intelligence as a latent factor/variable underlying fluid and crystallized intelligence. This is based on the extensive literature of classical g theory of intelligence [added in line 269~284].

Judd, N., Sauce, B., Wiedenhoeft, J., Tromp, J., Chaarani, B., Schliep, A., ... & Klingberg, T. (2020). Cognitive and brain development is independently influenced by socioeconomic status and polygenic scores for educational attainment. Proceedings of the National Academy of Sciences, 117(22), 12411-12418.

Martin, J., Hamshere, M. L., Stergiakouli, E., O'Donovan, M. C., & Thapar, A. (2015). Neurocognitive abilities in the general population and composite genetic risk scores for attention‐deficit hyperactivity disorder. Journal of Child Psychology and Psychiatry, 56(6), 648-656.

van Os, J., Linscott, R., Myin-Germeys, I., Delespaul, P., & Krabbendam, L. (2009). A systematic review and meta-analysis of the psychosis continuum: Evidence for a psychosis proneness–persistence–impairment model of psychotic disorder. Psychological Medicine, 39(2), 179-195. doi:10.1017/S0033291708003814

1. It is overall not clear when genetically and when self-reported information of ethnicity was used. This needs to be clearer throughout.

Thank you for mentioning this point. We only used genetically defined ethnicity, and we have not mentioned that we used self-reported ethnicity. Per your suggestion, we clarified that we used ‘genetic ethnicity’ throughout the paper.

1. The sentence starting line 253 is also unclear. How is schizophrenia PGS a 'more direct genetic predictor of PLE' and compared to what other measure?

Thank you for pointing this out. Please note that our adjustment (or sensitivity analyses) was based on the reported associations between PLEs and the risk for schizophrenia: schizophrenia PGS is associated with a cognitive deficit in psychosis patients (Shafee et al., 2018) and individuals at-risk of psychosis (He et al., 2021), and psychotic-like experiences (more so than PGS for psychotic-like experiences) (Karcher et al., 2018). We added these references for clarification [line 307~309]. We believe that because of the adjustment our results from the mixed linear model show the sensitivity and specificity of the association between cognitive phenotype PGS and PLEs.

He, Q., Jantac Mam-Lam-Fook, C., Chaignaud, J., Danset-Alexandre, C., Iftimovici, A., Gradels Hauguel, J., . . . Chaumette, B. (2021). Influence of polygenic risk scores for schizophrenia and resilience on the cognition of individuals at-risk for psychosis. Translational Psychiatry, 11(1). doi:10.1038/s41398-021-01624-z

Karcher, N. R., Paul, S. E., Johnson, E. C., Hatoum, A. S., Baranger, D. A. A., Agrawal, A., . . . Bogdan, R. (2021). Psychotic-like Experiences and Polygenic Liability in the Adolescent Brain Cognitive Development Study. Biological Psychiatry: Cognitive Neuroscience and Neuroimaging. doi:https://doi.org/10.1016/j.bpsc.2021.06.012

Shafee, R., Nanda, P., Padmanabhan, J. L., Tandon, N., Alliey-Rodriguez, N., Kalapurakkel, S., . . . Robinson, E. B. (2018). Polygenic risk for schizophrenia and measured domains of cognition in individuals with psychosis and controls. Translational Psychiatry, 8(1). doi:10.1038/s41398-018-0124-8

1. Please include a statement on the assumptions made when using the method used in this study and developed by Miao 2022, explain what evidence you have to support these assumptions and how this method, which I believe was developed for RCTs, can be applied to observational data.

We specified the assumptions for the causal inference method proposed by Miao et al. (2022) and why it is applicable to our study. Also, we noted that this novel method was developed to identify the causal effects of multiple treatment variables within non-randomized, observational data [line 309~319].

1. Some of the statements are potentially misleading. E.g., I would be very cautious to claim that the methods applied allowed the authors to estimate 'unbiased associations again potential (even unobserved) confounding variables'. There are many concerns such as selection bias, attrition, reverse causation, genetic confounding, etc that cannot be addressed satisfactorily using these data and methods.

Thank you for pointing this out. We deleted statements like ‘unbiased estimates’ and used expressions such as ‘adjustment for observed/unobserved confounding’ instead.

Nevertheless, please note that due to some limitations in the data (e.g., confounders), an analytic approach should be robust enough to handle potential violations of assumptions. This was the point we wanted to emphasize--In contrast to the majority of studies using the ABCD study, which employ simplistic GLM or conventional SEM with only latent variable modeling, our study provides less biased, thus more accurate, estimates through the use of sophisticated modeling for confounding effects (instead of simplistic GLM) and IGSCA (instead of conventional simplistic SEM). We hope our study may help improve our analytical approach in this field.

1. I would be equally cautious to claim that the ABCD study is representative. Please add information on the whole ABCD cohort to Table 1 and describe any relevance with respect to attrition effects or representativeness.

Thank you for highlighting this issue. We previously characterized the ABCD Study as representative of the US population, given its aim to ensure representativeness by recruiting from a broad range of school systems located near each of its 21 research sites, chosen for their geographic, demographic, and socioeconomic diversity. Using epidemiological strategies, a stratified probability sample of schools was selected for each site. This procedure took into account sex, race/ethnicity, socioeconomic status, and urbanicity to reduce potential sampling biases at the school level. Based on these strategies, previous research (e.g., Thompson et al., 2019; Zucker et al., 2018) has referred to the ABCD Study as ‘representative.’ However, we overlooked the fact that “not all 9-year-old and 10-year-old children in the United States had an equal chance of being invited to participate in the study,” and therefore, it should not be deemed fully representative of the US population (Compton et al., 2019). Heeding your suggestion, we have removed all descriptions of the ABCD Study being representative.

Compton, W. M., Dowling, G. J., & Garavan, H. (2019). Ensuring the Best Use of Data: The Adolescent Brain Cognitive Development Study. JAMA Pediatrics, 173(9), 809-810. doi:10.1001/jamapediatrics.2019.2081

Thompson, W. K., Barch, D. M., Bjork, J. M., Gonzalez, R., Nagel, B. J., Nixon, S. J., & Luciana, M. (2019). The structure of cognition in 9 and 10 year-old children and associations with problem behaviors: Findings from the ABCD study’s baseline neurocognitive battery. Developmental Cognitive Neuroscience, 36, 100606. doi:10.1016/j.dcn.2018.12.004

Zucker, R. A., Gonzalez, R., Feldstein Ewing, S. W., Paulus, M. P., Arroyo, J., Fuligni, A., . . . Wills, T. (2018). Assessment of culture and environment in the Adolescent Brain and Cognitive Development Study: Rationale, description of measures, and early data. Developmental Cognitive Neuroscience, 32, 107-120. doi:https://doi.org/10.1016/j.dcn.2018.03.004

1. The imputation methods need to be explained in more detail / more clearly. What concrete variables were included? Why was 50% of the sample excluded despite imputation? How similar is the study sample to the overall ABCD cohort - and to the US population in general (i.e., is this a representative dataset)?

Thank you for mentioning this point. We clarified the method and detailed processes of the imputation (e.g., R package VIM, number of missing observations for each study variables such as genotypes, follow-up observations, and positive environment) [Methods; line 167~176].

The final samples had significantly higher cognitive intelligence, parental education, family income, and family history of psychiatric disorders, lower Area Deprivation Index, percentage of individuals below -125% of the poverty level, and family’s financial adversity (p<0.05). As you have noted above, these results also show the limited representativeness of the data used in our study. We fully acknowledge that our study sample, as well as the overall ABCD cohort, is not representative of the US population in general.

1. There are a range of unclear statements (e.g., 'Supportive parenting and a positive school environment had the largest total impact on PLEs than genetic or environmental factors' - isn't parenting an environmental factor?).

Thank you for mentioning this point. We clarified seemingly vague expressions and unclear statements. We corrected the sentence you noted as ‘Supportive parenting and a positive school environment had the largest total impact on PLEs than any other genetic or environmental factors’ [line 57~58].

1. The authors' conclusion (that these findings have policy implications for improving school and family environmental) are not fully supported by the evidence. E.g., genetic effects were equally large.

Thank you for pointing this out. Our description should be clearer. Our models consistently show that the combined environmental effects of positive family/school environment, and family/neighborhood SES exceeds the genetic effects. We suggest that these findings may have policy implications for “improving the school and family environment and promoting local economic development” [line 62~64].

To clarify, we newly added “Despite the undeniable genetic influence on PLEs, when we combine the total effect sizes of neighborhood and family SES, as well as positive school environment and parenting behavior (∑▒〖|β|〗=0.2718~0.3242), they considerably surpass the total effect sizes of cognitive phenotypes PGSs (|β|=0.0359~0.0502)” [line 510~513]. Based on these results, we suggest that our findings hold potential policy implications for “preventative strategies that target residential environment, family SES, parenting, and schooling—a comprehensive approach that considers the entire ecosystem of children's lives—to enhance children's cognitive ability and mental health” in the Discussion [line 507~510].

Admittedly, our results do not directly demonstrate a causal effect wherein an intervention in the school or family environmental variables would necessarily lead to a significantly meaningful positive impact on a child's cognitive intelligence and mental health. We do not make such a claim in this paper. However, we anticipate that further integrative analyses akin to ours might help identify potential causal or prescriptive effects. We hope this perspective will be recognized as one of the contributions of our study. We leave the final decision to the discerning judgment of the editors and reviewers.

1. Many citations do not support the statements made and are sometimes used rather vaguely. For example, I believe Judd 2020 and Okbay 2022 did not use a PGS of cognitive capacity, but of educational attainment. Plomin 2018 and Harden 2020 are reviews, but the primary studies should be cited instead. Which reference exactly is supporting the statement that cognitive capacity PGS links to brain morphometry?

Thank you very much for your precise observations. We thoroughly checked all citations and updated the references for each statement.

We deleted Plomin & von Stumm (2018) and Harden & Koellinger (2020) and cited relevant original research articles (e.g., Lee et al., 2018; Okbay et al., 2022; Abdellaoui et al., 2022) instead. We also specified the references supporting the statement that educational attainment PGS links to brain morphometry (Judd et al., 2020; Karcher et al., 2021). As Okbay et al. (2022) used the PGS of cognitive intelligence (which presented the analyses results in their supplementary materials) as well as educational attainment, we decided to continue citing this reference [line 131~141].

1. Citations are formatted inconsistently.

We apologize for the inconsistency of the citation formatting. We formatted all citations in APA 7th style, using EndNote v20. We checked that all citations maintain consistency according to the reference style.

1. Re line 281, I believe effect sizes are 'up to twice as large', but not consistently twice as large as suggested in the text.

Thank you for mentioning this point. We corrected the sentence as ‘The effect sizes of EA PGS on children's PLEs were larger than those of CP PGS’ [line 342~343].

1. Please add to the results a short statement on what covariates these analyses were controlled for.

Thank you for giving us this comment. We added that we used sex, age, marital status, BMI, family history of psychiatric disorders, and ABCD research sites as covariates in the Results section [line 329~331].

1. Cho 2020 does not provide recommendations on FIT values (line 315). Please provide another reference and explain how these FIT values should be interpreted.

Thank you for mentioning this point. We added the correct reference for FIT values (Hwang, Cho, & Choo, 2021). We also added that the FIT values range from 0 to 1, and a larger FIT value indicates more variance of all variables is explained by the specified model (e.g., FIT=0.50 denotes that the model explains 50% of the total variance of all variables) [line 291~293].

1. Regarding Figure 2, please add factor loadings to this figure and explain what the difference between the hexagon and circular shapes are. Please also add the autocorrelations between the 3 PLE measures. I assume these were also modelled statistically, given the strong correlations between time points?

Figure 2B needs reworking.

It is unclear what the x-axis of Figure 2C represents. Proportion of R2 or effect size? SM table 2 provides key information, which should be added to Figure 2.

Thank you for pointing this out. We added factor loadings to the corrected figure (Figure 3A and 3B). We also added that the X-axis of Figure 3C represents standardized effect sizes.

1. I suggest adding units directly to Table 1, not in the legend. Was genetic or self-reported ethnicity used in this table? List age in years, not months?

Thank you for your suggestion. We added the units of age and family history of psychiatric disorders directly inside Table 1. We used genetic ethnicity in Table 1, as we only used genetic ethnicity (but not self-reported ethnicity) throughout our study. This is noted on the last row of Table 1. We listed age in chronological months, which is how each child’s age at each point of data collection is coded in the ABCD Study.

1. Please include exact p-values in Table 2.

Thank you for your suggestion. We highly appreciate the reviewer’s comment on the importance of showing exact p-values in the analysis results. Unfortunately, we cannot estimate the standard errors based on normal-theory approximations to obtain the exact p-values of our IGSCA model results. This is described in detail in the original paper of the IGSCA method (Hwang et al., 2021): “Like GSCA and GSCAM, IGSCA is also a nonparametric or distribution-free approach in the sense that it estimates parameters without recourse to distributional assumptions such as multivariate normality of indicators. As a trade-off of no reliance on distributional assumptions, it cannot estimate the standard errors of parameter estimates based on asymptotic (normal-theory) approximations. Instead, it utilizes the bootstrap method (Efron, 1979, 1982) to obtain the standard errors or confidence intervals of parameter estimates nonparametrically.”

Efron, B. (1979). Bootstrap methods: Another look at the jackknife. Annals of Statistics, 7, 1–26. http://dx.doi.org/10.1214/aos/1176344552

Efron, B. (1982). The jackknife, the bootstrap and other resampling plans. Philadelphia, PA: SIAM. http://dx.doi.org/10.1137/1.9781611970319

Hwang, H., Cho, G., Jung, K., Falk, C. F., Flake, J. K., Jin, M. J., & Lee, S. H. (2021). An approach to structural equation modeling with both factors and components: Integrated generalized structured component analysis. Psychological Methods, 26(3), 273-294. doi:10.1037/met0000336

1. There are way too many indigestible tables presented in the supplementary materials, which are also not referenced in the main manuscript.

We appreciate your insightful observation. As you rightly identified, we inadvertently failed to reference Table S2 in the main text. We have since corrected this omission in the Results section for the IGSCA (SEM) analysis [line 376]. The remainder of the supplementary tables (Table S1, S3~S7) have been appropriately cited in the main manuscript. We recognize that the quantity of tables provided in the supplementary materials is substantial. However, given the comprehensiveness and complexity of our analyses, these tables offer intricate results from each analysis. We deem these results, which include valuable findings from sensitivity analyses and confound testing, too significant to exclude from the supplementary materials. That said, we are open to, and would greatly welcome, any further suggestions to ensure clarity and ease of comprehension. Your guidance in this matter is highly valued.

1. Figure S1 is unclear, possibly due to the journal formatting. Is this one figure presented on two pages? Clarify which PGS is listed in Figure S1 and in any case, please add both PGSs.

Thank you for mentioning this point. Figure S1 presents two correlation matrices: the first one is the correlation matrix of component / factor variables in the IGSCA model and the second one is the that of observed variables used to construct the relevant component / factor variables in the IGSCA model. We noted each matrix as Figure S1-A and Figure S1-B. We also corrected the figure legend as “A. Correlation between all component / factor variables of the IGSCA model. B. Correlation between all observed variables used to construct the relevant component / factor variables in the IGSCA model.” Since Figure S1-A presents correlations between the components and latent factors, it lists a single PGS variable constructed from the CP PGS and EA PGS. On the other hand, Figure S1-B presents correlations between the observed variables. Thus, both CP PGS and EA PGS are listed in this correlation matrix.

Version 2

Abstract

Abstract

Background

In children, psychotic-like experiences (PLEs) are related to risk of psychosis, schizophrenia, and other mental disorders. Maladaptive cognitive functioning is a well-known risk factor and early marker for psychosis, schizophrenia, and other mental disorders. Since cognitive functioning is linked to various genetic and environmental factors during development, we hypothesize that it mediates the effects of those factors on childhood PLEs. Using large longitudinal data, we tested the relationships of genetic and environmental factors (such as familial and neighborhood environment) with cognitive intelligence and their relationships with current and future PLEs in children.

Methods

To estimate associations against potential confounding bias, we leveraged large-scale multimodal data of 6,602 children (aged 9-10 years old; 47.15% females; 5,211 European-ancestry) from the Adolescent Brain and Cognitive Development Study. Linear mixed model and a novel structural equation modeling (SEM) method that allows estimation of both components and factors were used to estimate the joint effects of cognitive phenotypes polygenic scores (PGSs), familial and neighborhood socioeconomic status (SES), and supportive environment on NIH Toolbox cognitive intelligence and PLEs. We adjusted for ethnicity (genetically defined), schizophrenia PGS, and additionally unobserved confounders (using computational confound modeling).

Results

We identified that lower cognitive intelligence and higher PLEs correlated significantly with several genetic and environmental variables: i.e., lower PGSs for cognitive phenotypes, lower familial SES, lower neighborhood SES, lower supportive parenting behavior, and lower positive school environment. In SEM, lower cognitive intelligence significantly mediated the genetic and environmental influences on higher PLEs (Indirect effects of PGS: β range=-0.0355∼ -0.0274; Family SES: β range=-0.0429∼ -0.0331; Neighborhood SES: β range=0.0126∼ 0.0164; Positive Environment: β range=-0.0039∼ -0.003). Supportive parenting and a positive school environment had the largest total impact on PLEs (β range=-0.152∼ -0.1316) than any other genetic or environmental factors.

Conclusions

Our results reveal the role of genetic and environmental factors on children’s PLEs via its negative impact on cognitive intelligence. Our findings have policy implications in that improving the school and family environment and promoting local economic development might be a way to enhance cognitive and mental health in children.

Key Points

While existing research shows the association between cognitive decline and PLEs, the genetic and environmental pathways to cognitive intelligence and psychotic risk in children remain unclear.

We identified the significant role of genetic and environmental factors (family, neighborhood, and school) on children’s PLEs via a negative impact on cognitive intelligence.

Leveraging large samples with multimodal longitudinal data and advanced computational modeling for adjustment of observed/unobserved confounding bias, our results underscore the importance of incorporating socioeconomic policies into children’s cognitive and mental health programs.

Introduction

Childhood is the critical developmental period in human life. Cognitive intelligence and mental health in this period significantly impact key life outcomes at later ages, including academic performance, economic productivity, physical health, intelligence, and psychopathology (Shonkoff, 2012; Walker et al., 2021). Literature shows the significant impact of social adversities on cognitive ability and mental health in early childhood. Lower family socioeconomic status (SES), particularly household income, is linked to lower neurocognitive ability and higher risk of psychopathology in childhood (Hair et al., 2015; Noble et al., 2015; Peverill et al., 2021; Tomasi & Volkow, 2021; Weissman et al., 2018).

Additional to family SES, the importance of neighborhood social environment on children’s neurocognitive ability has been also emphasized (Gard et al., 2021; Tooley et al., 2020). Adverse neighborhood environment, such as the percent of families below poverty line, low education levels, and exposure to violence, is associated with lower cognitive performance and a greater risk for psychosis in children (Butler et al., 2018; Karcher, Schiffman, et al., 2021; Rakesh et al., 2021; Taylor et al., 2020). Conversely, as protective factors against familial and neighborhood socioeconomic challenges, supportive parenting (Brody et al., 2017; Brody et al., 2019; Holmes et al., 2018; Luby et al., 2012; Luby et al., 2016) and positive school environment (Gard et al., 2021; Piccolo et al., 2019; Rakesh et al., 2021) have been highlighted to improve child cognition and mental health.

Psychotic-like experiences (PLEs), which are prevalent in childhood and often precede the onset of schizophrenia (Howes & Murray, 2014), indicate not only the risk of psychosis (van der Steen et al., 2019; Van Os & Reininghaus, 2016) but also point towards the potential for other psychopathologies including mood, anxiety, and substance disorders (Poulton et al., 2000; van der Steen et al., 2019). PLEs are linked to a reduced cognitive intelligence (Cannon et al., 2002; Kelleher & Cannon, 2011) and show a stronger association with environmental risk factors during childhood than other internalizing/externalizing symptoms (Karcher, Schiffman, et al., 2021).

Maladaptive cognitive intelligence may precede the manifestation of psychotic symptoms (Cannon et al., 2000; Keefe et al., 2006; Reichenberg et al., 2005). An integrated theoretical model of schizophrenia posits that anomalous neurodevelopment, influenced by genetic factors, combined with disrupted cognitive processes resulting from socioenvironmental adversity, may eventually give rise to the positive symptoms of schizophrenia, relevant to PLEs (Garety et al., 2001; Howes & Murray, 2014). Family studies show a decline in cognitive intelligence preceding psychotic symptoms is related to genetic risk (Cosway et al., 2000; Curtis et al., 2001). In more recent studies these associations have led to the model positing that cognitive intelligence mediates the genetic risk for psychopathology and PLEs (Karcher, Paul, et al., 2021; Pat et al., 2021). Moreover, there is evidence of a shared genetic component between psychotic disorders and cognitive phenotypes, such as educational attainment (Escott-Price et al., 2020; Le Hellard et al., 2016).

To minimize potential bias in the estimates of environmental effects, it is crucial to adjust for genetic confounding (Sariaslan et al., 2016), given the substantial genetic influence on intelligence (Bouchard & McGue, 1981; Deary et al., 2006; Plomin & Spinath, 2004) and PLEs (Bentall & Fernyhough, 2008; Maxwell et al., 2023). Recent advances in genetics have led to the development of the polygenic score (PGS) approach: a computational method to estimate the genetic loading for a complex trait using statistical associations of each single-nucleotide polymorphism (SNP) identified by genome-wide association studies (GWAS) (Choi et al., 2020). Particularly, PGS for two related but distinct phenotypes—cognitive performance and educational attainment—holds significant importance. As the two most frequently used proxies of cognitive intelligence in genetic studies, PGSs for cognitive performance and educational attainment are positively correlated with intelligence, educational attainment, income, self-rated health, and height (Judd et al., 2020; Lee et al., 2018; Okbay et al., 2022; Selzam et al., 2019). Furthermore, the PGS for educational attainment is associated with a wide range of biological and social outcomes, including brain morphometry (Judd et al., 2020; Karcher, Paul, et al., 2021), psychopathologies such as PLEs, autism, depression, Alzheimer’s disease, neuroticism (Karcher, Paul, et al., 2021; Okbay et al., 2022), cognitive decline (Joo et al., 2022; Karcher, Paul, et al., 2021; Ritchie et al., 2020), BMI, time spent watching television, geographic residence (Abdellaoui et al., 2022), and wealth inequality (Barth et al., 2020). Similar to the terms used in prior research, we will collectively refer to these two PGSs of focus as ‘cognitive phenotypes PGSs’ throughout this paper (Joo et al., 2022; Okbay et al., 2022; Selzam et al., 2019). An important gap in the literature is the lack of integrated assessment of the effects of genetic and environmental factors at multiple levels (e.g., familial vs neighborhood) to dissect the genetic and environmental effects underlying abnormal cognitive intelligence and the PLEs. Addressing this with large multimodal data will allow for a more complete understanding of the factors related to the development of PLEs.

In this study, we systematically explore the longitudinal trajectories of genetic and environmental influences on PLEs, mediated through cognitive intelligence. Toward this goal, we firstly assess the associations of cognitive phenotype PGSs, family and neighborhood SES, and supportive environment with children’s cognitive intelligence and longitudinally measured PLEs. To maintain robustness of our assessment, we employed statistical and computational approaches to carefully consider potential confounding. We then test the mediating effect of cognitive intelligence on the relationship between genetic and environmental factors and PLEs. Our investigation traces these effects from the baseline and through the 1-year and 2-year follow-ups, providing a nuanced understanding on the role of cognitive phenotype polygenic scores, family socioeconomic status, neighborhood socioeconomic status, and positive family and school environments in shaping PLEs in children aged 9-10 years.

Materials and Methods

Study Participants

We used the multi-modal genetic and environmental data of 11,878 preadolescent children aged 9-10 years old collected from 21 research sites of the Adolescent Brain Cognitive Development (ABCD) Study, one of the largest longitudinal studies for children’s neurodevelopment in the United States. We analyzed the baseline, first year, and second year follow-up datasets included in ABCD Release 4.0, downloaded on January 25, 2022. After k-nearest neighbor imputation of missing values of covariates (categorical variables: sex, genetic ethnicity, marital status of the caregiver, ABCD research sites; continuous variables: age, BMI, family history of psychiatric disorders; 4.67% of total observations imputed) using the R package VIM (Kowarik & Templ, 2016), we removed participants with missing data on study variables (missing genotype: N=3,260; follow-up observations: N=1,180; neighborhood information: N=694; cognitive intelligence tests: N=126; PLEs: N=5; positive environment: N=11). The final samples included 6,602 multiethnic children, which comprised 890 of African ancestry (13.48%), 91 of East Asian ancestry (1.38%), 5,211 of European ancestry (78.93%), 229 of Native American ancestry (3.47%), and 181 not specified (2.74%).

Data

NIH Toolbox Cognitive Performance

Children’s neurocognitive abilities were assessed using the NIH Toolbox Cognitive Battery, which has seven cognitive instruments for examining executive function, episodic memory, language abilities, processing speed, working memory, and attention (Thompson et al., 2019). We utilized baseline observations of uncorrected composite scores of fluid intelligence (Dimensional Change Card Sort Task, Flanker Test, Picture Sequence Memory Test, List Sorting Working Memory Test), crystallized intelligence (Picture Vocabulary Task and Oral Reading Recognition Test), and total intelligence (all seven instruments) provided in the ABCD Study dataset.

Psychotic-Like Experiences (PLEs)

Baseline and 1- and 2-year follow-up of PLEs were measured using the children’s responses to the Prodromal Questionnaire-Brief Child Version. In line with previous research (Karcher et al., 2018; Karcher et al., 2020; Karcher, Schiffman, et al., 2021), we computed Total Score and Distress Score, each indicating the number of psychotic symptoms and levels of total distress. Considering self-reports and parent-reports of psychopathology may differ (Achenbach, 2006), we additionally used parent-rated PLEs derived from four items of the Child Behavior Checklist according to previous studies (Karcher et al., 2018; Karcher et al., 2020; Karcher, Schiffman, et al., 2021). Self-reported PLEs and parent-reported PLEs had significant positive correlation (Pearson’s correlation of baseline year: r = 0.095∼0.0989, p<0.0001; 1-year follow-up: r = 0.1322∼0.1327, p<0.0001; 2-year follow-up: r = 0.1569∼0.1632, p<0.0001).

Polygenic Scores (PGS)

To investigate the aggregated effect of genetic components, we estimated PGS of two representative cognitive phenotypes for each participant: educational attainment (EA) and cognitive performance (CP) (Choi et al., 2020). We used the summary statistics released from a GWAS (Lee et al., 2018) of European-descent individuals for EA (n=1,131,881) and CP (n=257,841). EA was measured as the years of schooling; CP was measured as the respondent’s score on cognitive ability assessments. To construct PGS of schizophrenia for sensitivity analyses, we used the summary statistics from the multiple GWAS of European sample (n=65,967; Ruderfer et al. (2018)) and East Asian sample (n=58,140; Lam et al. (2019)). We applied PRS-CSx, a high-dimensional Bayesian regression framework that improves cross-population prediction via continuous shrinkage prior to SNP effect sizes (Ge et al., 2019) (For details, see Appendix S1). The two PGSs for cognitive phenotypes had a positive significant correlation (Pearson’s correlation: r =0.4331, p<0.0001).

Family-, Neighborhood-, and School-level Environment

We assessed children’s family-level SES with family income, parental education, and family’s financial adversity based on parent self-reporting (Karcher et al., 2020; Taylor et al., 2020; Tomasi & Volkow, 2021). Higher family income and parental education and lower family’s financial adversity denote a higher family SES.

Neighborhood-level SES was assessed using the Area Deprivation Index (ADI), the percentage of individuals below -125% of the poverty level (henceforth “poverty”), and years of residence, which were associated with PLEs in prior research (Karcher, Schiffman, et al., 2021). Higher values of ADI and poverty and fewer years of residence indicate a lower neighborhood SES.

Based on existing literature (Karcher, Schiffman, et al., 2021; Rakesh et al., 2021), we measured the level of positive parenting behavior and positive school environment to assess the effect of positive family and school environment on each individual.

Statistical Modeling

In this study, we employ linear mixed models and a novel structural equation modelling (SEM) method to examine the longitudinal trajectories of genetic and environmental influences on PLEs mediated by cognitive intelligence. We specifically investigate the mediating role of cognitive intelligence within the impacts of cognitive phenotype polygenic scores (PGSs), high family socioeconomic status (SES), low neighborhood SES, and positive family and school environments on PLEs. These influences are examined across three periods of PLEs observations: baseline, 1-year follow-up, and 2-year follow-up (Figure 1).

Study diagram for longitudinal trajectories of genetic and environmental influences on PLEs through cognitive intelligence.

This study examines the mediating role of cognitive intelligence within the effects of cognitive phenotype PGSs, high family SES, low neighborhood SES, and positive family and school environments on PLEs observed at baseline, 1-year follow-up, and 2-year follow-up in children aged 9-10 years. Both direct and indirect effects, as well as total effects, were evaluated for statistical significance.

Linear Mixed Models

Using linear mixed models to minimize potential population stratification from different household environments and geographical locations (Choi et al., 2020), we analyzed the genetic and environmental effects on cognitive intelligence and PLEs. Key variables of interest were PGS, family SES, neighborhood SES, and positive environment. To avoid multicollinearity, CP and EA PGSs were used separately in two other models. As a random factor, the variable indicating ABCD research sites was used. Variables within each model had no signs of multicollinearity (Fox & Monette, 1992) (generalized variance inflation factor < 2.0 for every variable in all models). The child’s sex, age, genetic ethnicity, BMI, marital status of the caregiver, and family history of psychiatric disorders were included as covariates in the model (Karcher, Schiffman, et al., 2021). All continuous variables were standardized (z-scaled) beforehand to get standardized estimates, and the analyses were conducted with the lme4 package in R version 4.1.2. Throughout this paper, threshold for statistical significance was set at p<0.05, with correction for multiple comparisons using the false discovery rate. 95% bootstrapped confidence intervals were obtained with 5,000 iterations.

Path Modeling

To test the plausibility of whether cognitive intelligence may mediate the association between genetic and environmental factors and PLEs, we used an up-to-date SEM method, integrated generalized structured component analysis (IGSCA) (Hwang, Cho, Jung, et al., 2021). This approach is suited to our study using the multi-modal genetic and environmental variables in that it estimates models with both factors and components as statistical proxies for the constructs. Unlike standard SEM with latent factors, the IGSCA method allows for the use of composite indicators (i.e., components)—defined as a weighted sum of observed variables—as constructs in the model, which can more effectively control bias in estimation compared to the standard SEM. During estimation, the IGSCA determines weights of each observed variable in such a way as to maximize the variances of all endogenous indicators and components.

We considered path-analytic relationships among the six key constructs: cognitive phenotypes PGSs, family SES, neighborhood SES, positive family and school environment, general intelligence, and PLEs. Prior research indicates that the relevant observed variables of the PGSs, family SES, neighborhood SES, positive family and school environment, and PLEs tend not to have a common latent factor and are consequently evaluated as a composite index in analyses. Notably, investigations carried out by Judd et al. (2020) and Martin et al. (2015) utilized composite indicators to examine the genetic influence on educational attainment and ADHD. Moreover, socioenvironmental influences are often treated as composite indicators as highlighted in Judd et al. (2020). When considering the psychosis continuum, studies like that of van Os et al. (2009) postulate that psychotic disorders are likely underpinned by a multiplicity of background factors rather than a single common factor. This perspective is substantiated by a multitude of prior research that deploys composite indices for the measurement of psychotic symptoms. For these reasons, we statistically represented these constructs as the weighted sums of their observed variables or components (Cho et al., 2021). On the other hand, we represented general intelligence as a common factor that determines the underlying covariance pattern of fluid and crystallized intelligence, based on the classical g theory of intelligence (Jensen, 1998; Spearman, 1904).

The IGSCA model included the same covariates used in the linear mixed model as well as the ABCD research site as an additional covariate. We applied GSCA Pro 1.1 (Hwang, Cho, & Choo, 2021) to fit the IGSCA model to the data and checked the model’s goodness-of-fit index (GFI) (Jöreskog & Sörbom, 1986), standardized root mean square residual (SRMR), and total variance of all indicators and components explained (FIT) to assess its overall goodness-of-fit. Ranging from 0 to 1, a larger FIT value indicates more variance of all variables is explained by the specified model (e.g., FIT=0.50 denotes that the model explains 50% of the total variance of all variables) (Hwang, Cho, & Choo, 2021). The rules-of-thumb cutoff criteria in IGSCA is GFI≥0.93 and SRMR<0.08 for an acceptable fit (Cho et al., 2020). Finally, we conducted conditional process analyses to investigate further the indirect and total effects of the constructs in the model. We obtained 95% confidence intervals based on 5,000 bootstrap samples to test the statistical significance of parameter estimates.

Sensitivity Analyses

To ensure robustness of the main analyses results, we conducted multiple sensitivity analyses. As the European-descent-based GWAS was used for constructing PGS, we reran the main analyses using participants of European ancestry (n=5,211) to adjust for ethnic confounding. Next, we tested effects of gene x environment interactions on cognitive intelligence and PLEs, respectively. We also assessed whether the effects of cognitive phenotypes PGS in the linear mixed model are significant after adjusting for schizophrenia PGS, given that prior studies reported the significant association of schizophrenia PGS and cognitive deficit in psychosis patients (Shafee et al., 2018) and individuals at-risk of psychosis (He et al., 2021). Lastly, we adjusted for unobserved confounding bias in the linear mixed model, using a recently developed framework for causal inference based on null treatments approach (Miao et al., 2022). Designed to discern causal effects from multiple treatment variables within non-randomized, observational data, the null treatments approach hinges on the assumption that no fewer than half of the confounded treatments exert no causal influence on the outcome. It circumvents the need for prior knowledge regarding which treatments are null and eliminates the necessity for independence among treatments.

Given our model’s inclusion of numerous treatment variables with shared variances due to the presence of unobserved confounders (Abdellaoui et al., 2022; Okbay et al., 2022)—including cognitive phenotypes PGS, family and neighborhood SES, positive family and school environments— we opted to employ this method.

Results

Demographics

Table 1 presents the demographic characteristics of the final samples. For multiethnic subjects (main analyses, n=6,602), 47.15% were female, and the parents of 70.21% were married. In European ancestry samples (sensitivity analyses, n=5,211), 46.71% were female, and the parents of 77.47% were married. Children of European ancestry showed significantly different marital status (p<0.0001), lower BMI (p<0.0001) and family history of psychiatric disorders (p<0.0001) compared to children of other genetic ancestries. Our linear mixed model and IGSCA analyses were adjusted using sex, age, marital status, BMI, family history of psychiatric disorders, and ABCD research sites as covariates.

Genetic influence on cognitive phenotypes correlates positively with intelligence and negatively with PLEs

As shown in Figure 2, higher PGSs of cognitive capacity phenotypes correlated significantly with higher intelligence (CP PGS: β=0.1113∼0.1793; EA PGS: β=0.0699∼0.1567). While CP PGS was associated only with lower baseline year Distress Score PLEs (β=-0.0323), EA PGS was associated with lower baseline year and follow-up PLEs of all measures (baseline: β=-0.0518∼-0.0519; 1-year: β=-0.0423∼-0.043; 2-year: β=- 0.036∼ -0.0463). No significant correlations were found between CP PGS and follow-up PLEs (p>0.05). The effects of EA PGS were larger on baseline year PLEs than follow-up PLEs. The effect sizes of EA PGS on children’s PLEs were larger than those of CP PGS (EA PGS: β=-0.036∼-0.0519; CP PGS: β=-0.0323). (Table S1).

Linear models testing genetic, socioeconomic, and environmental factors associated with cognitive intelligence and PLEs.

Standardized coefficients of a linear mixed model with CP PGS (A, B) and EA PGS (C, D). The analyses included 6,602 samples of multiethnicity. Cognitive intelligence and PLEs correlated with the PGSs, residential disadvantage, positive environment, and family SES in the opposite directions. Error bars indicate 95% bootstrapped confidence intervals with 5,000 iterations. CP and EA denote cognitive performance and education attainment, respectively; PGS, polygenic scores; SES, socioeconomic status; PLEs, psychotic-like experiences; ADI, Area Deprivation Index; Poverty, percentage of individuals below -125% of the poverty level; Years, years of residence. (*p-FDR<0.05)

Family and neighborhood SES correlates positively with intelligence and negatively with PLEs

Parental education associated positively with all types of intelligence (β=0.0699∼0.1745) and negatively with baseline year Total and Distress Score PLEs (β=- 0.0528∼-0.043), 1-year follow-up PLEs (β=-0.0538∼-0.0449), and 2-year follow-up PLEs (β=-0.0459∼-0.0389). Family income correlated positively with intelligence (β=0.0723∼0.1365) and negatively with 2-year follow-up parent-rated PLEs (β=-0.0503∼ - 0.0502). Family’s financial disadvantage correlated negatively with baseline year parent-rated PLEs (β=0.0726∼0.0728), 1-year follow-up PLEs of all types (β=0.0307∼0.0577), and 2-year follow-up PLEs of all types (β=0.0461∼0.0581).

The ADI correlated significantly negatively with all types of intelligence (β=- 0.0684∼-0.054). Additionally, a higher ADI correlated significantly with higher baseline year PLEs (β=0.0587∼0.0914), 1-year follow-up PLEs (β=0.0523∼0.0613), and 2-year follow-up PLEs (β=0.0397∼0.0449).

We found no significant associations of poverty with any of the target variables (p> 0.05). Years of residence correlated significantly with crystallized intelligence (β=0.035∼0.0372) and baseline year Total Score PLEs (β=-0.029∼-0.0273). (Table S1).

Positive family and school environments correlate positively with intelligence and negatively with the influence of PLEs

Positive parenting behaviors showed significant negative correlations with baseline year PLEs (β=-0.0702∼-0.0419), 1-year follow-up PLEs (β=-0.0588∼-0.0397), and 2-year follow-up PLEs (β=-0.0623∼-0.0356) (Figure 2). Positive school environment was associated positively with total intelligence (β=0.0353∼0.0397) and fluid intelligence (β=0.0514∼0.0545) and negatively with all three measures of baseline year PLEs (β=-0.1193∼-0.0468), 1-year follow-up PLEs (β=-0.1078∼-0.04), and 2-year follow-up PLEs (β=-0.1068∼-0.0586) (Table S1).

Structural Equation Modeling-Integrated Generalized Structured Component Analysis

The IGSCA model showed that intelligence mediated the effects of genes and environments on the risk for psychosis (PLEs) (Figure 3 and Table 2). Estimated factor loadings of latent factor variable and weights of component variables are presented in Table S2. The model showed a good model fit with a GFI of 0.9735, SRMR of 0.0359, and FIT value of 0.4912 (Cho et al., 2020). Intelligence was under significant direct influences of the cognitive phenotypes PGS (β=0.2427), family SES (β=0.2932), neighborhood SES (β=- 0.1121), and positive environment (β=0.0268). Family SES and positive environment had significant negative direct effects on PLEs of all years. Cognitive phenotypes PGS and neighborhood SES showed no significant direct effects on any of the PLEs (p>0.05).

Pathway diagram of the structural equation model of gene-environment-intelligence-PLEs pathway: Cognitive phenotypes PGSs, high family SES, low neighborhood SES, positive environment, intelligence, and PLEs. A. Standardized path coefficients are indicated on each path as direct effect estimates (significance level *p<0.05). B. Indirect pathways to PLEs via intelligence were significant for polygenic scores, high family SES, low neighborhood SES, and positive environment, indicating the significant mediating role of intelligence. C. Relative effect sizes of direct and indirect pathways within the total effects on PLEs. The standardized effect sizes of direct pathways are colored within each bar. In A-B, child sex, genetic ethnicity, BMI, marital status, family history of psychiatric disorders, and ABCD research sites were included as covariates. CP PGS and EA PGS denote polygenic scores of cognitive performance and education attainment, respectively; SES, socioeconomic status; PLEs, psychotic-like experiences; Crystallized and Fluid, crystallized and fluid intelligence; ADI, Area Deprivation Index; Poverty, percentage of individuals below -125% of the poverty level; Years, years of residence. Note: * indicates a statistically significant parameter estimate at α=0.05.

Intelligence significantly mediated the effects of the PGS, family and neighborhood SES, and positive environment on PLEs of all years: the PGS (baseline year: β=-0.035; 1-year: β=- 0.0355; 2-year: β=-0.0274), family SES (baseline year: β=-0.0423; 1-year: β=--0.0429; 2- year: β=-0.0331), neighborhood SES (baseline year: β=0.0162; 1-year: β=0.0164; 2-year: β=0.0126), and positive environment (baseline year: β=-0.0039; 1-year: β=-0.0039; 2-year: β=-0.003).

For all observed years, positive environment had largest total effects on PLEs (baseline year: β=-0.152; 1-year: β=-0.1316; 2-year: β=-0.1364), followed by family SES (baseline year: β=-0.1216; 1-year: β=-0.1119; 2-year: β=-0.1164), neighborhood SES (baseline year: β=0.0504; 1-year: β=0.0329; 2-year: β=0.0192), and PGS (baseline year: β=- 0.0498; 1-year: β=-0.036; 2-year: β=-0.0365). The total effects of each indicator on PLEs were significant except for those of neighborhood SES (p>0.05).

Sensitivity Analyses

As sensitivity analyses, we reran our main analyses with adjustment for ethnic confounding, schizophrenia PGS, and unobserved confounding, respectively. Results of linear mixed model and IGSCA analyses were consistent with European ancestry participants (Table S3, S4). Interactions between PGS and positive parenting behavior had significant positive effects on 2-year follow-up Total and Distress Score PLEs (β=0.0343∼0.053) (Table S5). Unlike CP PGS and EA PGS, schizophrenia PGS did not have significant associations with any of the PLEs (p>0.05). Associations of cognitive capacitive PGS, family and neighborhood SES, and positive environment were similar after adjusting for schizophrenia PGS (See Table S6 for details). Lastly, when adjusting for unobserved confounders, CP and EA PGS, positive parenting behavior, and positive school environment remained significantly correlated with intelligence and PLEs of all years (Table S7).

Discussion

This study investigated the relationships of the genetic and environmental influences on the development of intelligence and the PLEs in youth, leveraging genetic data from the large epidemiological samples and a multi-level environmental (socioeconomic) data. Our results support the model that genetic factors (PGS for cognitive phenotypes), socioeconomic conditions, and family and school environments may influence cognitive intelligence in children, and this impact may lead to the individual variability of the current and future PLEs in children. Our analysis with integrated data shows the contributions of genetic and environmental factors, respectively, to cognitive and mental wellness in children, and further provides policy implications to improve them.

Our SEM shows that cognitive intelligence mediates the environmental and genetic influence on the current and future PLEs. The environmental factors (family SES, neighborhood SES, and positive parenting and schooling) and PGS of cognitive phenotypes exhibit significant indirect effects via cognitive intelligence on PLEs. Prior studies identifying the mediation of cognitive intelligence focused on either genetic or environmental factors alone (Karcher, Paul, et al., 2021; Lewis et al., 2020). Our study advances this by demonstrating the specific effects of genetic and environmental factors while controlling for each other. Considering the directions and magnitudes of the effects, though the effects of PGS remain significant, aggregated effects of environmental factors account for much greater degrees on PLEs (potential policy implications is discussed below). The indirect pathway of PGS & environment-intelligence-PLEs aligns with the studies with older clinical samples showing cognitive deficits as a precursor for the onset of psychosis and schizophrenia and their genetic and environmental factors (Eastvold et al., 2007; Fett et al., 2020; Lewis et al., 2020; Vorstman et al., 2015).

Our results of cognitive intelligence mediating the genetic and environmental effects on PLEs may be related to several potential mechanisms. Children raised in higher family SES may have sufficient nutrition and cognitive stimulants, whereas children living in deprived neighborhoods may be exposed to higher rates of crime, air pollution, and substance abuse (Lewis et al., 2020; Marshall et al., 2020; Taylor et al., 2020; Tomasi & Volkow, 2021). Environmental enrichment may be associated with longer periods of neural plasticity (e.g. myelination, maturation of brain circuitry), leading to higher cognitive ability and lower risk of mental disorders like PLEs (Tooley et al., 2021). This may be further linked to the cognitive reserve theory. The theory suggests that genetic influence for cognitive phenotypes and environmental enrichment promotes more efficient, flexible brain networks, which may lead to greater resilience against psychopathology (Stern, 2009). Indeed, prior clinical studies show the linkage between cognitive reserve and psychosis (Amoretti et al., 2018; Leeson et al., 2011).

Our results indicate that genetic influences on cognitive phenotypes are significantly linked to PLEs. While schizophrenia PGS showed no significant associations with PLEs, PGSs for CP and EA were strongly correlated with PLEs (baseline year, 1-year follow-up, and 2-year follow-up). These associations were robust after adjustment for ethnic confounding and unobserved confounders. Cognitive phenotypes PGS generally show higher predictive performance than PGS of any other traits (Lee et al., 2018; Okbay et al., 2022; Plomin & von Stumm, 2018). Genetic variants associated with CP and EA are related to complex traits across the life span, including neuroticism, depressive symptoms, smoking in adulthood, cognitive decline at a later age (Joo et al., 2022), risk for Alzheimer’s disease (Lee et al., 2018; Okbay et al., 2022), brain volume, area, and thickness, as well as psychotic disorders (Karcher, Paul, et al., 2021). Prior gene expression studies suggest that polygenic signals for schizophrenia, bipolar disorder, and EA are significantly enriched in the central nervous system, particularly the cerebellum (Finucane et al., 2018). Our findings emphasize the importance of cognitive phenotypes PGS as a biomarker which not only implicates cognitive traits but also exhibits genetic overlap with the PLEs.

The differing magnitudes of the PGS impact between EA and CP warrant attention. The effects of the EA PGS on the PLEs of all years were 160.68%∼371.67% larger than those of CP PGS. This discrepancy may result from that the larger sample size of EA GWAS than that of CP GWAS. Alternatively, the discrepancies in effect sizes may suggest different genetic compositions between EA and CP. Recent literature documents that more than half of the polygenic signal for EA is related to noncognitive and social skills required for successful educational attainment (Demange et al., 2021), whereas CP may rather be linked to cognitive skills. This observation also supports the well-established relationships of the EA PGS with socioeconomic and life-course outcomes (e.g., social mobility (Belsky et al., 2018), voter turnout (Aarøe et al., 2020), BMI, income, time spent watching television, geographic residence (Abdellaoui et al., 2022), and wealth inequality (Barth et al., 2020), which may be influenced by unobserved environmental factors (Young et al., 2019). In our analysis, the utilization of two PGSs a more comprehensive evaluation, contributing to an estimation of the genetic and environmental factors that attempted to minimize confounding bias.

Furthermore, the significant effects of cognitive phenotypes PGS on cognitive intelligence (β=0.0699∼0.1793) remained robust and similar in magnitude after adjusting for genetic ethnicity (β=0.0754∼0.1866) and other (unobserved) confounding (β=0.0546∼0.1776). As we controlled for family-, neighborhood-, and school-level environmental factors and unobserved confounders, our results may be interpreted as significant genetic influences on individual’s cognitive intelligence. This interpretation is supported by a recent study (Isungset et al., 2022): Despite of the socioeconomic differences in Norway (a typical social democratic welfare state) and the US (a typical liberal welfare state), the magnitudes of genetic influence on cognitive intelligence were similar (Norway: β=0.18; US: β=0.17). Cognitive phenotypes PGS is an important genetic factor across the nations and societies. Therefore, analyses omitting the genetic influence may be subject to overestimation of the socioeconomic impact (Plomin & von Stumm, 2018; Sariaslan et al., 2016).

This study shows that a high SES and positive environment, particularly positive parenting behavior and school environment, is associated with higher intelligence and a lower risk for PLEs in children. While prior research has emphasized the dominant role of family SES (e.g. family income) (Tomasi & Volkow, 2021), our SEM analyses(IGSCA) showed that positive environmental factors such as supportive parenting and schooling have a greater impact on children’s PLEs. Specifically, the effect sizes were the highest in supportive family and school environment, followed by family and neighborhood SES. Even after adjusting for genetic ethnicity and unobserved confounders, the strong associations of positive parenting and schooling with higher intelligence and fewer PLEs remained significant. These findings suggest that interventions that target positive family and school environments may be particularly effective. Recent research supports this notion, showing that interventions that promote supportive parenting and inclusive school environments can improve neurocognitive ability, academic performance, and decrease risk behaviors such as drinking and emotional eating (Brody et al., 2017; Brody et al., 2019; Holmes et al., 2018).

Moreover, our results showed that positive parenting and schooling in baseline year were associated not only with baseline year PLEs but also with PLEs 1-2 years later. This is in line with prior research showing that intervention focused on parenting behavior and school environment have long-lasting positive effects that extend into adulthood and even across generations (Cunha & Heckman, 2007; Hill et al., 2020).

Our findings highlight the potential for preventative strategies that target residential environment, family SES, parenting, and schooling—a comprehensive approach that considers the entire ecosystem of children’s lives—to enhance children’s cognitive ability and mental health. Despite the undeniable genetic influence on PLEs, when we combine the total effect sizes of neighborhood and family SES, as well as positive school environment and parenting behavior (∑ |P| = 0.2718∼0.3242), they considerably surpass the total effect sizes of cognitive phenotypes PGSs (|P| = 0.0359∼0.0502). A holistic and quantitative approach that takes into account the comprehensive ecosystem of family, school, and residential environments may ensure treatment effects and efficient use of resources (Cree et al., 2018; Garner & Yogman, 2021; Shonkoff, 2012). For example, the Health Impact in 5 Years Initiative of the US Centers for Disease Control and Prevention (CDC, 2018) includes 14 evidence-based interventions, such as providing school-based prevention programs, public transportation, home improvement loans, and earned income tax credits, to tackle the social determinants of public health. Our study strengthens the idea that an interdisciplinary science-driven, coordinated approach to intervening in the select environmental factors may allow practical improvements in child development, particularly in those who are at a disadvantage.

Our study has some limitations. First, due to data availability constraints in the ABCD study, we only utilized baseline observations for NIH Toolbox cognitive intelligence, and we could not test whether PLEs might be a mediator of intelligence. Second, the generalizability of our findings may be limited since most of the participants included in our analysis are from European ancestry. Although the ABCD Study aimed to achieve its representativeness by recruiting from an array of school systems located around each of the 21 research sites, chosen for their diversity in geography, demographics, and socioeconomic status, it is not fully representative of the US population (Compton et al., 2019). Third, the duration of the follow-up period utilized in this study is relatively short (1-year and 2-year follow-up), which may limit the interpretability of our findings for understanding cognitive and psychiatric development during later childhood. Future research could potentially benefit from employing longer follow-up periods, as more follow-up observations are being collected in the ABCD Study. Fourth, while we used a wide range of statistical methods to adjust for confounding bias from observed and unobserved variables (e.g., genetic ethnicity), we did not account for other types of potential bias such as sample selection bias. Fifth, despite a number of causal inference methods used in this study, the ABCD Study is a non-randomized dataset. Given the observational nature of the ABCD Study, interpreting our results as actual causality requires more caution. Finally, we did not include all important environmental variables, such as air pollution (Marshall et al., 2020) and social capital (Krabbendam, 2005), which are not collected in the ABCD study.

In conclusion, our study provides potential pathways of genetic factors of cognitive phenotypes and environmental factors of family, school, and neighborhood to cognitive and mental wellness in children. Our findings underscore the importance of a comprehensive approach that considers both biological and socioeconomic features in promoting young children’s cognitive ability and mental health. Given the importance of child development, it requires joint efforts from multiple disciplines.

Funding and Disclosure

This work was supported by the New Faculty Startup Fund from Seoul National University, a Research grant from Basic Science Research Program through the National Research Foundation of Korea (NRF) funded by the Ministry of Education, South Korea (2021R1I1A1A01054995), and Institute of Information & communications Technology Planning & Evaluation (IITP) grant funded by the Korea government (MSIT) [NO.2021-0- 01343, Artificial Intelligence Graduate School Program (Seoul National University)]. The authors have nothing to disclose.

Author Contributions

JP, EL, and GC conducted the main analyses. HH, YYJ, and JC supervised the study. JP, EL preprocessed data. JP, EL, GC, HH performed statistical analyses. EL, BK, GK, and YYJ constructed PGS. JP, EL, GC, YYJ, and JC drafted and critically reviewed the manuscript. All authors read and approved the final draft.

Data Availability

All codes used in this study can be found at https://github.com/Transconnectome/Intell_PLE_Pathway. Due to ABCD Study’s policy in data sharing, we provide a synthetic dataset instead of real observations. This synthetic dataset was generated with conditional GAN (Xu et al., 2019) to imitate the data structure of our final study samples. After automatic hyperparameter optimization with Optuna (Akiba et al., 2019), the synthetic dataset showed overall quality score of 84.15%. Note that analyses results from the synthetic data may not be 100% identical to the results presented in this paper, due to the differences between synthetic vs original dataset.

eLife Assessment

This study presents a useful inventory of the joint effects of genetic and environmental factors on psychotic-like experiences, and identifies cognitive ability as a potential underlying mediating pathway. The data were analyzed using solid and validated methodology based on a large, multi-center dataset. The claim that these findings are of relevance to psychosis risk and have implications for policy changes are partially supported by the results.

Reviewer #1 (Public Review):

This study by Park et al. describes an interesting approach to disentangle gene-environment pathways to cognitive development and psychotic-like experiences (PLEs) in children. They have used data from the ABCD (Adolescent Brain Cognitive Development) study and have included phenotypes polygenic scores (PGS) of educational attainment (EA) and cognition, environmental exposure data, cognitive performance data and self-reported PLEs. The study has several strengths, including its large sample size, interesting approach and comprehensive statistical model,

One remaining concern is the authors' conflation of PLEs and schizophrenia. They stated, for example, that it is necessary to adjust for schizophrenia PGS. Even though studies have found a statistical relationship between schizophrenia PGS and PLEs, this relationship is not very strong (although statistically significant) and other studies have found no relationship. Similarly, having PLEs increases the risk of developing psychosis, but that does not necessarily mean that this risk is substantial or specific. I think this needs more nuance in the manuscript and the term 'schizophrenia' should be used sparsely and very carefully as the paper has focused on PLEs.

Reviewer #2 (Public Review):

This paper tried to assess the link between genetic and environmental factors on psychotic-like experiences, and the potential mediation through cognitive ability. This study was based on data from the ABCD cohort, including 6,602 children aged 9-10y. The authors report a mediating effect, suggesting that cognitive ability is a key mediating pathway in the link between several genetic and environmental (risk and protective) factors on PLEs.

Strengths of the methods

The authors use a wide range of validated (genetic, self- and parent-reported, as well as cognitive) measures in a large dataset with a 2-year follow-up period. The statistical methods have potential to address key limitations of previous research.

Weaknesses of the methods

The methodological advantage of the method (Integrated generalized structured component analysis, IGSCA) over the standard method (Structural equation modeling, SEM) is not fully clear.

Not all methods are fully explained (how genetic components were derived; how cognition was assessed in Lee et al., 2018).

Not the largest or most recent GWAS (Genome-wide association studies) were used to generate PGS.

Strengths of the results

The authors included a comprehensive array of analyses.

Weaknesses of the results

Some factor loadings presented in Figure 3 seem counterintuitive/inconsistent.

Supplementary tables are difficult to assess. Unclear significance statement / p-values in Table 2.

Appraisal

The authors suggest that their findings provide evidence for policy reforms (e.g., targeting residential environment, family SES (social economic status), parenting, and schooling).

Impact

Immediate impact is limited given the short follow-up period (2y), possibly concerns for selection bias and attrition in the data, and some methodological concerns. The authors are transparent about most of these limitations.

Author Response

The following is the authors’ response to the previous reviews.

Our comments on the initial eLife assessment

“This study presents a useful inventory of the joint effects of genetic and environmental factors on psychotic-like experiences, and identifies cognitive ability as a potential underlying mediating pathway. The data were analyzed using solid and validated methodology based on a large, multi-center dataset. The claim that these findings are of relevance to psychosis risk and have implications for policy changes are partially supported by the results”

We sincerely appreciate the editor and reviewers for their valuable feedback and their willingness to accommodate our perspectives in the first revision. In this revision, the comments from the reviewers have allowed us to further improve our manuscript. Regarding the eLife assessment, we would like to discuss two points.

Firstly, regarding your point of our “findings are of relevance to psychosis risk…partially supported…”, we want to address that our study is closely related to psychosis risk. Childhood psychotic-like experiences (PLEs) are closely linked to psychotic risk and have been shown to increase the risk of general psychopathology, as mentioned in our Introduction and Discussion.

The reviewers asked for clearer differentiation between PLEs and schizophrenia, which we incorporated in this revision (line 100~111; line 419~430). So, this revised version now clearly points out that findings are relevant primarily to psychosis risk, and only partially relevant to schizophrenia risk.

Secondly, regarding “…implications for policy changes are partially supported…”, we have revised our study’s social contribution more clearly and specifically. Incorporating the comments, we have revised that our study offers an insight to the future studies by showing the importance of integrative approaches, considering multi-factorial neurocognition and psychopathology ranging from genes to environment (line 503~512), rather than offers direct policy implications.

Our collaboration with eLife and the reviewers has proven satisfactory and enriching. The community, coupled with the innovative system and culture established around eLife, has significantly advanced the progression of scientific research. We are privileged to contribute to this endeavor.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

I am happy with the revisions provided by the authors and I think most of my concerns have been addressed satisfactorily. One remaining concern is the authors' conflation of PLEs and schizophrenia. They stated, for example, that it is necessary to adjust for schizophrenia PGS. Even though studies have found a statistical relationship between schizophrenia PGS and PLEs, this relationship is not very strong (although statistically significant) and other studies have found no relationship. Similarly, having PLEs increases the risk of developing psychosis, but that does not necessarily mean that this risk is substantial or specific. I think this needs more nuance in the manuscript and the term 'schizophrenia' should be used sparsely and very carefully as the paper has focused on PLEs. Otherwise, great work on the revisions, thank you.

Thank you for your comment on the use of PLEs and schizophrenia. We clearly understand the differences between the two and we made relevant corrections throughout the manuscript. In particular, we added that PLEs are not a direct predictor of schizophrenia and corrected any expressions that may imply that PLEs are closely related to schizophrenia in the Introduction.

“Psychotic-like experiences (PLEs), which are prevalent in childhood, indicate the risk of psychosis (van der Steen et al., 2019; Van Os & Reininghaus, 2016). Although they are not a direct precursor of schizophrenia, children reporting PLEs in ages of 9-11 years are at higher risk of psychotic disorders in adulthood (Kelleher & Cannon, 2011; Poulton et al., 2000). PLEs also point towards the potential for other psychopathologies including mood, anxiety, and substance disorders (van der Steen et al., 2019), are linked to deficits in cognitive intelligence (Cannon et al., 2002; Kelleher & Cannon, 2011) and show a stronger association with environmental risk factors during childhood than other internalizing/externalizing symptoms (Karcher, Schiffman, et al., 2021).

Maladaptive cognitive intelligence may act as a mediator for the effects of genetic and environmental risks on the manifestation of psychotic symptoms (Cannon et al., 2000; Keefe et al., 2006; Reichenberg et al., 2005).” (line 100~111)

We also revised any expressions that could be perceived as implying relevance to schizophrenia in the Discussion. “Prior research identifying the mediation of cognitive intelligence focused on either genetic (Karcher, Paul, et al., 2021) or environmental factors (Lewis et al., 2020) alone. Studies with older clinical samples have shown that cognitive deficit may be a precursor for the onset of psychotic disorders (Eastvold et al., 2007; Fett et al., 2020; Vorstman et al., 2015). Our study advances this by demonstrating the integrated effects of genetic and environmental factors on PLEs through the cognitive intelligence in 9-11 years old children. Such comprehensive analysis contributes to assessing the relative importance of various factors influencing children's cognition and mental health, and it can aid future studies designed for identifying health policy implications. Considering the directions and magnitudes of the effects, though the effects of PGS remain significant, aggregated effects of environmental factors account for much greater degrees on PLEs.” (line 419~430)

Reviewer #2 (Recommendations For The Authors):

I thank the authors for addressing most of my comments. I feel the manuscript has already greatly improved.

I have a few more comments.

1. Although I did not make this comment, I find the authors' reply to the following comment by Reviewer #1 unclear: Original comment 'I like that the assessment of CP (cognitive performance) and self-reports PLEs is of good quality. However, I was wondering which 4 items from the parent-reported CBCL (Child Behavior Checklist) were used and how did they correlate with the child-reported PLEs? And how was distress taken into account in the child self-reported PLEs measurement? Which PLEs measures were used?'

The authors' response refers to correlation coefficients, but I think Reviewer #1's inquiry was on more than these correlations.

Thank you for your concern. We think that this comment was referring to our previous manuscript submitted elsewhere. In our initial submission to eLife, we already added the details about the four items from the parent-reported CBCL and how distress was considered in the child self-reported PLEs measurement (Appendix S1, page 48).

1. Regarding the authors' reply that they have 'standardized the use of 'cognitive capacity' - I do not understand what this means. How exactly was this term standardized? In fact, I can find the term 'cognitive capacity' only once and it seemed to have been deleted from the manuscript. This is fine, but it doesn't clearly align with the statement that this term has been standardized.

We apologize for causing such confusion. What we meant was that throughout our revised manuscript, we used the term “cognitive phenotypes” instead of “cognitive capacity”.

1. Regarding my initial comment that 'it needs to be described how cognitive performance was defined in Lee 2018.' - I believe this is still not clarified. The authors write 'CP was measured as the respondent's score on cognitive ability assessments', but it remains unclear what exactly these assessments were.

Thank you for pointing this out. We added that “CP, measured as the respondent's score on cognitive ability assessments of general cognitive function and verbal-numerical reasoning, was assessed in participants from the COGENT consortium and the UK Biobank” (line 204~206).

1. Regarding the authors' reply to my comment 'In the 'Path Modeling' section, please explain what 'factors and components' concretely refer to. How is this different from a standard SEM with latent factors?'

I can see that the authors explained 'components' (=the weighted sum of observed variables), but please also add what you mean by 'factors' - and how these are different from 'components' (line 284). Furthermore, I don't think it is correct that SEMs can only model latent factors, but not components (=measured variables). I also cannot see how using a weighted sum of observed variables controls more effectively for bias in estimation than latent factors. However, even though I do have some knowledge on this method, I'm not an expert and would appreciate the authors, other reviewer and/or editor to weigh in on this point.

Thank you for pointing this out. We added that latent factors are indirectly measured indicators that explain the covariance among observed variables (line 263~271). We also added that standard SEM method using latent factors assumes that observed variables within each construct share a common underlying factor, but if this assumption is not met, then the standard SEM method cannot effectively control for biases. This is the reason why the IGSCA method, which addresses this limitation by allowing for use of both composite and latent factors as constructs.

“Standard SEM using latent factors (i.e., indirectly measured indicators that explain the covariance among observed variables) to represent indicators such as PGS or family SES relies on the assumption that observed variables within each construct share a common underlying factor. If this assumption is violated, standard SEM cannot effectively control for estimation biases. The IGSCA method addresses this limitation by allowing for the use of composite indicators (i.e., components)—defined as a weighted sum of observed variables—as constructs in the model, more effectively controlling bias in estimation compared to the standard SEM. During estimation, the IGSCA determines weights of each observed variable in such a way as to maximize the variances of all endogenous indicators and components.” (line 263~271)

1. I overall disagree with the authors' following statement 'It has been suggested from prior studies that these variables (PGS, family SES, neighborhood SES, positive family and school environment, and PLEs) are less likely to share a common factor', but I appreciate the authors' argument.

Thank you for your comment. To make clarify our statement in the manuscript, we changed the sentence to “Considering that the observed variables of the PGSs, family SES, neighborhood SES, positive family and school environment, and PLEs are evaluated as a composite index by prior research, the IGSCA method can mitigate bias more effectively by representing these constructs as components” (line 274~277).

1. Regarding 'genetic ethnicity': please describe your methods on how this was defined.

Genetic ethnicity was defined as the genetic ancestry of participants, which is included as one of observations in the original ABCD Study data. To avoid further confusion, we corrected ‘genetic ethnicity’ to ‘genetic ancestry’ throughout the manuscript.

1. Regarding 'a more direct genetic predictor of PLEs' - I still don't understand what the contrast is here. More direct than what else?

The description was unclear; we removed it from our manuscript.

1. Regarding the factor loadings in Figure 3: I don't understand how deprivation loads positively on 'low neighborhood SES', but poverty loads negatively. Shouldn't they both show the same direction of effect/loading on neighbourhood SES, while 'years of residency' should show the opposite direction (i.e., deprivation and poverty = risk, while years of residency = protective)? Are these unexpected loadings?

The authors did not yet respond to this point: 'Please also add the autocorrelations between the 3 PLE measures. I assume these were also modelled statistically, given the strong correlations between time points?' Were these correlations not modelled? Why not?

Figure 3B is still unclear. Was intelligence included here? What is the difference between Figure 3A and B? The legend suggests that 3B shows the indirect effects, but figure 3B looks like a direct effect, while 3A seem to show the indirect effect.

The reviewer’s confusion resulted from our incorrect description. The factor loadings of low neighborhood SES were marked incorrectly. The loading for ‘years of residence’ and ‘poverty’ should be switched: -0.3648 for ‘years of residence’ and +0.877 for ‘poverty’. This was a mistake when we were applying factor loadings in the Figure. We thank you for pointing this out.

We apologize for missing your point on autocorrelation. Adding autocorrelations between the three PLEs is unrelated to our research goal. In this paper, we investigated how genetic and environmental factors explain the variations in PLEs between participants, regardless of changes over time. Since we used PLEs of multiple follow-ups to ensure that the results are robust irrespective of the timing of PLE measurements, taking autocorrelation into account is not necessary.

The decision to add autocorrelation, which involves using the outcome variable at time (t-1) as a predictor for the outcome variable at time t, depends on the research focus. If your interest lies in explaining inter-individual variation in the rate of change in PLEs over a one-year period, then autocorrelation should be controlled for (typically, predictors measured at different time points are used in such cases). However, this was not the focus of this paper, which is why we did not apply autocorrelation in the SEM analysis.

We apologize for the confusion between Figure 3A and 3B. To clarify, we added titles in the figure images as “Direct effects” and “Indirect effects”. We also changed the legend as well.

“A. Direct pathways from PGS, high family SES, low neighborhood SES, and positive environment to cognitive intelligence and PLEs. Standardized path coefficients are indicated on each path as direct effect estimates (significance level *p<0.05). B. Indirect pathways to PLEs via intelligence were significant for polygenic scores, high family SES, low neighborhood SES, and positive environment, indicating the significant mediating role of intelligence.” (line 968~973)

Figure 3A shows direct effects: i.e., the coefficients of paths from PGS, family SES, neighborhood SES, and positive environment to intelligence and PLEs, as well as the coefficient of paths from intelligence to PLEs. This is why Figure 3A shows colored arrows starting from PGS, family and neighborhood SES, and positive environment towards intelligence and PLEs, as well as the arrows from intelligence to PLEs. On the other hand, in Figure 3B, the colored arrows staring from PGS, family and neighborhood SES, and positive environment goes through intelligence, and heads towards PLEs. This was meant to show that the indirect effects shown in Figure 3B indicate the specific effects of PGS, family SES, neighborhood SES, and positive environment on PLEs mediated by intelligence.

In short, Figure 3 can be seen as a diagram drawn from Table 2: direct effects of the genetic and environmental variables on intelligence and PLEs, and direct effects of intelligence on PLEs are shown in Figure 3A; indirect effects of genetic and environmental variables on PLEs mediated by intelligence are shown in Figure 3B.

1. Regarding Supporting Information tables: to make these more digestible, I suggest using Excel and adding one table per sheet with a clear title and legend, indicating what each table shows. For example, Table S1 has 9(?) different subsections, all called the same (Linear Mixed Model: Multiethnic). It is not clear how each subsection differs from the others. Separate tables in separate excel sheets might be easier.

Also, I think two decimal points might be good enough, enhancing readability of these tables.

Thank you for your suggestion. We moved the supplementary tables into an external Excel file, with each sheet showing different tables, as well as titles, legends, and clear subsections.

1. Regarding reporting exact p-values in Table 2: I don't understand. At the moment, categorical significance statements are reported. Were these not based on exact p-values (or how else was it decided if a finding was significant at a 0.05 (?) significance level).

Either remove the significance column completely (as p-values cannot be estimated due to non-normality) or specify exactly/clarify what this column shows and this was derived.

We apologize for the confusion. In Table 2, we checked the significance of each path using 95% confidence intervals with 5,000 bootstrapping iterations. Since 95% confidence intervals that does not include zero is equivalent to p-values below 0.05 significance level, we believe this is an appropriate alternative for reporting the significance of each path in the SEM model.

We specified the reason why we were not able to calculate exact p-values (clean copy: line 299~303). “As a trade-off for obtaining robust nonparametric estimates without distributional assumptions for normality, the IGSCA method does not return exact p-values (Hwang, Cho, Jung, et al., 2021). As a reasonable alternative, we obtained 95% confidence intervals based on 5,000 bootstrap samples to test the statistical significance of parameter estimates.”

Version 3

Abstract

Abstract

Background

In children, psychotic-like experiences (PLEs) are related to risk of psychosis, schizophrenia, and other mental disorders. Maladaptive cognitive functioning is a well-known risk factor and early marker for psychosis, schizophrenia, and other mental disorders. Since cognitive functioning is linked to various genetic and environmental factors during development, we hypothesize that it mediates the effects of those factors on childhood PLEs. Using large longitudinal data, we tested the relationships of genetic and environmental factors (such as familial and neighborhood environment) with cognitive intelligence and their relationships with current and future PLEs in children.

Methods

To estimate associations against potential confounding bias, we leveraged large-scale multimodal data of 6,602 children (aged 9-10 years old; 47.15% females; 5,211 European-ancestry) from the Adolescent Brain and Cognitive Development Study. Linear mixed model and a novel structural equation modeling (SEM) method that allows estimation of both components and factors were used to estimate the joint effects of cognitive phenotypes polygenic scores (PGSs), familial and neighborhood socioeconomic status (SES), and supportive environment on NIH Toolbox cognitive intelligence and PLEs. We adjusted for ethnicity (genetically defined), schizophrenia PGS, and additionally unobserved confounders (using computational confound modeling).

Results

We identified that lower cognitive intelligence and higher PLEs correlated significantly with several genetic and environmental variables: i.e., lower PGSs for cognitive phenotypes, lower familial SES, lower neighborhood SES, lower supportive parenting behavior, and lower positive school environment. In SEM, lower cognitive intelligence significantly mediated the genetic and environmental influences on higher PLEs (Indirect effects of PGS: β range=-0.0355~ -0.0274; Family SES: β range=-0.0429~ -0.0331; Neighborhood SES: β range=0.0126~ 0.0164; Positive Environment: β range=-0.0039~ -0.003). Supportive parenting and a positive school environment had the largest total impact on PLEs (β range=-0.152~ -0.1316) than any other genetic or environmental factors.

Conclusions

Our results reveal the role of genetic and environmental factors on children’s PLEs via its negative impact on cognitive intelligence. Our findings have policy implications in that improving the school and family environment and promoting local economic development might be a way to enhance cognitive and mental health in children.

Key Points

While existing research shows the association between cognitive decline and PLEs, the genetic and environmental pathways to cognitive intelligence and psychotic risk in children remain unclear.

We identified the significant role of genetic and environmental factors (family, neighborhood, and school) on children’s PLEs via a negative impact on cognitive intelligence.

Leveraging large samples with multimodal longitudinal data and advanced computational modeling for adjustment of observed/unobserved confounding bias, our results underscore the importance of incorporating socioeconomic policies into children’s cognitive and mental health programs.

Introduction

Childhood is the critical developmental period in human life. Cognitive intelligence and mental health in this period significantly impact key life outcomes at later ages, including academic performance, economic productivity, physical health, intelligence, and psychopathology (Shonkoff, 2012; Walker et al., 2021). Literature shows the significant impact of social adversities on cognitive ability and mental health in early childhood. Lower family socioeconomic status (SES), particularly household income, is linked to lower neurocognitive ability and higher risk of psychopathology in childhood (Hair et al., 2015; Noble et al., 2015; Peverill et al., 2021; Tomasi & Volkow, 2021; Weissman et al., 2018).

Additional to family SES, the importance of neighborhood social environment on children’s neurocognitive ability has been also emphasized (Gard et al., 2021; Tooley et al., 2020). Adverse neighborhood environment, such as the percent of families below poverty line, low education levels, and exposure to violence, is associated with lower cognitive performance and a greater risk for psychosis in children (Butler et al., 2018; Karcher, Schiffman, et al., 2021; Rakesh et al., 2021; Taylor et al., 2020). Conversely, as protective factors against familial and neighborhood socioeconomic challenges, supportive parenting (Brody et al., 2017; Brody et al., 2019; Holmes et al., 2018; Luby et al., 2012; Luby et al., 2016) and positive school environment (Gard et al., 2021; Piccolo et al., 2019; Rakesh et al., 2021) have been highlighted to improve child cognition and mental health.

Psychotic-like experiences (PLEs), which are prevalent in childhood, indicate the risk of psychosis (van der Steen et al., 2019; Van Os & Reininghaus, 2016). Although they are not a direct precursor of schizophrenia, children reporting PLEs in ages of 9-11 years are at higher risk of psychotic disorders in adulthood (Kelleher & Cannon, 2011; Poulton et al., 2000). PLEs also point towards the potential for other psychopathologies including mood, anxiety, and substance disorders (van der Steen et al., 2019), are linked to deficits in cognitive intelligence (Cannon et al., 2002; Kelleher & Cannon, 2011) and show a stronger association with environmental risk factors during childhood than other internalizing/externalizing symptoms (Karcher, Schiffman, et al., 2021).

Maladaptive cognitive intelligence may act as a mediator for the effects of genetic and environmental risks on the manifestation of psychotic symptoms (Cannon et al., 2000; Keefe et al., 2006; Reichenberg et al., 2005). Abnormal neurodevelopment, influenced by genetic factors, combined with disrupted cognitive processes resulting from socioenvironmental adversity, may eventually give rise to the positive symptoms of schizophrenia, relevant to PLEs (Garety et al., 2001; Howes & Murray, 2014). Family studies show a decline in cognitive intelligence preceding psychotic symptoms is related to genetic risk (Cosway et al., 2000; Curtis et al., 2001). In more recent studies these associations have led to the model positing that cognitive intelligence mediates the genetic risk for psychopathology and PLEs (Karcher, Paul, et al., 2021; Pat et al., 2021).

To minimize potential bias in the estimates of environmental effects, it is crucial to adjust for genetic confounding (Sariaslan et al., 2016), given the substantial genetic influence on intelligence (Bouchard & McGue, 1981; Deary et al., 2006; Plomin & Spinath, 2004) and PLEs (Bentall & Fernyhough, 2008; Maxwell et al., 2023). Recent advances in genetics have led to the development of the polygenic score (PGS) approach: a computational method to estimate the genetic loading for a complex trait using statistical associations of each single-nucleotide polymorphism (SNP) identified by genome-wide association studies (GWAS) (Choi et al., 2020). Particularly, PGS for two related but distinct phenotypes—cognitive performance and educational attainment—holds significant importance. As the two most frequently used proxies of cognitive intelligence in genetic studies, PGSs for cognitive performance and educational attainment are positively correlated with intelligence, educational attainment, income, self-rated health, and height (Judd et al., 2020; Lee et al., 2018; Okbay et al., 2022; Selzam et al., 2019). Furthermore, the PGS for educational attainment is associated with a wide range of biological and social outcomes, including brain morphometry (Judd et al., 2020; Karcher, Paul, et al., 2021), psychopathologies such as PLEs, autism, depression, Alzheimer’s disease, neuroticism (Karcher, Paul, et al., 2021; Okbay et al., 2022), cognitive decline (Joo et al., 2022; Karcher, Paul, et al., 2021; Ritchie et al., 2020), BMI, time spent watching television, geographic residence (Abdellaoui et al., 2022), and wealth inequality (Barth et al., 2020). Similar to the terms used in prior research, we will collectively refer to these two PGSs of focus as ‘cognitive phenotypes PGSs’ throughout this paper (Joo et al., 2022; Okbay et al., 2022; Selzam et al., 2019). An important gap in the literature is the lack of integrated assessment of the effects of genetic and environmental factors at multiple levels (e.g., familial vs neighborhood) to dissect the genetic and environmental effects underlying abnormal cognitive intelligence and the PLEs. Addressing this with large multimodal data will allow for a more complete understanding of the factors related to the development of PLEs.

In this study, we systematically explore the longitudinal trajectories of genetic and environmental influences on PLEs, mediated through cognitive intelligence. Toward this goal, we firstly assess the associations of cognitive phenotype PGSs, family and neighborhood SES, and supportive environment with children’s cognitive intelligence and longitudinally measured PLEs. To maintain robustness of our assessment, we employed statistical and computational approaches to carefully consider potential confounding. We then test the mediating effect of cognitive intelligence on the relationship between genetic and environmental factors and PLEs. Our investigation traces these effects from the baseline and through the 1-year and 2-year follow-ups, providing a nuanced understanding on the role of cognitive phenotype polygenic scores, family socioeconomic status, neighborhood socioeconomic status, and positive family and school environments in shaping PLEs in children aged 9-10 years.

Materials and Methods

Study Participants

We used the multi-modal genetic and environmental data of 11,878 preadolescent children aged 9-10 years old collected from 21 research sites of the Adolescent Brain Cognitive Development (ABCD) Study, one of the largest longitudinal studies for children’s neurodevelopment in the United States. We analyzed the baseline, first year, and second year follow-up datasets included in ABCD Release 4.0, downloaded on January 25, 2022. After k-nearest neighbor imputation of missing values of covariates (categorical variables: sex, genetic ancestry, marital status of the caregiver, ABCD research sites; continuous variables: age, BMI, family history of psychiatric disorders; 4.67% of total observations imputed) using the R package VIM (Kowarik & Templ, 2016), we removed participants with missing data on study variables (missing genotype: N=3,260; follow-up observations: N=1,180; neighborhood information: N=694; cognitive intelligence tests: N=126; PLEs: N=5; positive environment: N=11). The final samples included 6,602 multiethnic children, which comprised 890 of African ancestry (13.48%), 91 of East Asian ancestry (1.38%), 5,211 of European ancestry (78.93%), 229 of Native American ancestry (3.47%), and 181 not specified (2.74%).

Data

NIH Toolbox Cognitive Performance

Children’s neurocognitive abilities were assessed using the NIH Toolbox Cognitive Battery, which has seven cognitive instruments for examining executive function, episodic memory, language abilities, processing speed, working memory, and attention (Thompson et al., 2019). We utilized baseline observations of uncorrected composite scores of fluid intelligence (Dimensional Change Card Sort Task, Flanker Test, Picture Sequence Memory Test, List Sorting Working Memory Test), crystallized intelligence (Picture Vocabulary Task and Oral Reading Recognition Test), and total intelligence (all seven instruments) provided in the ABCD Study dataset.

Psychotic-Like Experiences (PLEs)

Baseline and 1- and 2-year follow-up of PLEs were measured using the children’s responses to the Prodromal Questionnaire-Brief Child Version. In line with previous research (Karcher et al., 2018; Karcher et al., 2020; Karcher, Schiffman, et al., 2021), we computed Total Score and Distress Score, each indicating the number of psychotic symptoms and levels of total distress. Considering self-reports and parent-reports of psychopathology may differ (Achenbach, 2006), we additionally used parent-rated PLEs derived from four items of the Child Behavior Checklist according to previous studies (Karcher et al., 2018; Karcher et al., 2020; Karcher, Schiffman, et al., 2021). Self-reported PLEs and parent-reported PLEs had significant positive correlation (Pearson’s correlation of baseline year: r = 0.095~0.0989, p<0.0001; 1-year follow-up: r = 0.1322~0.1327, p<0.0001; 2-year follow-up: r = 0.1569~0.1632, p<0.0001).

Polygenic Scores (PGS)

To investigate the aggregated effect of genetic components, we estimated PGS of two representative cognitive phenotypes for each participant: educational attainment (EA) and cognitive performance (CP) (Choi et al., 2020). We used the summary statistics released from a GWAS (Lee et al., 2018) of European-descent individuals for EA (n=1,131,881) and CP (n=257,841). EA was measured as the years of schooling; CP, measured as the respondent’s score on cognitive ability assessments of general cognitive function and verbal-numerical reasoning, was assessed in participants from the COGENT consortium and the UK Biobank. To construct PGS of schizophrenia for sensitivity analyses, we used the summary statistics from the multiple GWAS of European sample (n=65,967; Ruderfer et al. (2018)) and East Asian sample (n=58,140; Lam et al. (2019)). We applied PRS-CSx, a high-dimensional Bayesian regression framework that improves cross-population prediction via continuous shrinkage prior to SNP effect sizes (Ge et al., 2019) (For details, see Appendix S1). The two PGSs for cognitive phenotypes had a positive significant correlation (Pearson’s correlation: r =0.4331, p<0.0001).

Family-, Neighborhood-, and School-level Environment

We assessed children’s family-level SES with family income, parental education, and family’s financial adversity based on parent self-reporting (Karcher et al., 2020; Taylor et al., 2020; Tomasi & Volkow, 2021). Higher family income and parental education and lower family’s financial adversity denote a higher family SES.

Neighborhood-level SES was assessed using the Area Deprivation Index (ADI), the percentage of individuals below −125% of the poverty level (henceforth “poverty”), and years of residence, which were associated with PLEs in prior research (Karcher, Schiffman, et al., 2021). Higher values of ADI and poverty and fewer years of residence indicate a lower neighborhood SES.

Based on existing literature (Karcher, Schiffman, et al., 2021; Rakesh et al., 2021), we measured the level of positive parenting behavior and positive school environment to assess the effect of positive family and school environment on each individual.

Statistical Modeling

In this study, we employ linear mixed models and a novel structural equation modelling (SEM) method to examine the longitudinal trajectories of genetic and environmental influences on PLEs mediated by cognitive intelligence. We specifically investigate the mediating role of cognitive intelligence within the impacts of cognitive phenotype polygenic scores (PGSs), high family socioeconomic status (SES), low neighborhood SES, and positive family and school environments on PLEs. These influences are examined across three periods of PLEs observations: baseline, 1-year follow-up, and 2-year follow-up (Figure 1).

Study diagram for longitudinal trajectories of genetic and environmental influences on PLEs through cognitive intelligence.

This study examines the mediating role of cognitive intelligence within the effects of cognitive phenotype PGSs, high family SES, low neighborhood SES, and positive family and school environments on PLEs observed at baseline, 1-year follow-up, and 2-year follow-up in children aged 9-10 years. Both direct and indirect effects, as well as total effects, were evaluated for statistical significance.

Linear Mixed Models

Using linear mixed models to minimize potential population stratification from different household environments and geographical locations (Choi et al., 2020), we analyzed the genetic and environmental effects on cognitive intelligence and PLEs. Key variables of interest were PGS, family SES, neighborhood SES, and positive environment. To avoid multicollinearity, CP and EA PGSs were used separately in two other models. As a random factor, the variable indicating ABCD research sites was used. Variables within each model had no signs of multicollinearity (Fox & Monette, 1992) (generalized variance inflation factor < 2.0 for every variable in all models). The child’s sex, age, genetic ancestry, BMI, marital status of the caregiver, and family history of psychiatric disorders were included as covariates in the model (Karcher, Schiffman, et al., 2021). All continuous variables were standardized (z-scaled) beforehand to get standardized estimates, and the analyses were conducted with the lme4 package in R version 4.1.2. Throughout this paper, threshold for statistical significance was set at p<0.05, with correction for multiple comparisons using the false discovery rate. 95% bootstrapped confidence intervals were obtained with 5,000 iterations.

Path Modeling

To test the plausibility of whether cognitive intelligence may mediate the association between genetic and environmental factors and PLEs, we used an up-to-date SEM method, integrated generalized structured component analysis (IGSCA) (Hwang, Cho, Jung, et al., 2021). This approach is suited to our study using the multi-modal genetic and environmental variables in that it estimates models with both factors and components as statistical proxies for the constructs.

Standard SEM using latent factors (i.e., indirectly measured indicators that explain the covariance among observed variables) to represent indicators such as PGS or family SES relies on the assumption that observed variables within each construct share a common underlying factor. If this assumption is violated, standard SEM cannot effectively control for estimation biases. The IGSCA method addresses this limitation by allowing for the use of composite indicators (i.e., components)—defined as a weighted sum of observed variables— as constructs in the model, more effectively controlling bias in estimation compared to the standard SEM. During estimation, the IGSCA determines weights of each observed variable in such a way as to maximize the variances of all endogenous indicators and components.

We assessed path-analytic relationships among the six key constructs: cognitive phenotypes PGSs, family SES, neighborhood SES, positive family and school environment, general intelligence, and PLEs. Considering that the observed variables of the PGSs, family SES, neighborhood SES, positive family and school environment, and PLEs are evaluated as a composite index by prior research, the IGSCA method can mitigate bias more effectively by representing these constructs as components. Notably, investigations carried out by Judd et al. (2020) and Martin et al. (2015) utilized composite indicators to examine the genetic influence on educational attainment and ADHD. Moreover, socioenvironmental influences are often treated as composite indicators as highlighted in Judd et al. (2020). When considering the psychosis continuum, studies like that of van Os et al. (2009) postulate that psychotic disorders are likely underpinned by a multiplicity of background factors rather than a single common factor. This perspective is substantiated by a multitude of prior research that deploys composite indices for the measurement of psychotic symptoms. For these reasons, we statistically represented these constructs as the weighted sums of their observed variables or components (Cho et al., 2021). On the other hand, we represented general intelligence as a common factor that determines the underlying covariance pattern of fluid and crystallized intelligence, based on the classical g theory of intelligence (Jensen, 1998; Spearman, 1904).

The IGSCA model included the same covariates used in the linear mixed model as well as the ABCD research site as an additional covariate. We applied GSCA Pro 1.1 (Hwang, Cho, & Choo, 2021) to fit the IGSCA model to the data and checked the model’s goodness-of-fit index (GFI) (Jöreskog & Sörbom, 1986), standardized root mean square residual (SRMR), and total variance of all indicators and components explained (FIT) to assess its overall goodness-of-fit. Ranging from 0 to 1, a larger FIT value indicates more variance of all variables is explained by the specified model (e.g., FIT=0.50 denotes that the model explains 50% of the total variance of all variables) (Hwang, Cho, & Choo, 2021). The rules-of-thumb cutoff criteria in IGSCA is GFI≥0.93 and SRMR≤0.08 for an acceptable fit (Cho et al., 2020). Finally, we conducted conditional process analyses to investigate further the indirect and total effects of the constructs in the model. As a trade-off for obtaining robust nonparametric estimates without distributional assumptions for normality, the IGSCA method does not return exact p-values (Hwang, Cho, Jung, et al., 2021). As a reasonable alternative, we obtained 95% confidence intervals based on 5,000 bootstrap samples to test the statistical significance of parameter estimates.

Sensitivity Analyses

To ensure robustness of the main analyses results, we conducted multiple sensitivity analyses. As the European-descent-based GWAS was used for constructing PGS, we reran the main analyses using participants of European ancestry (n=5,211) to adjust for ethnic confounding. Next, we tested effects of gene x environment interactions on cognitive intelligence and PLEs, respectively. We also tested the effects of cognitive phenotypes PGS adjusting for schizophrenia PGS, given the association of schizophrenia PGS and cognitive deficit in psychosis patients (Shafee et al., 2018) and individuals at-risk of psychosis (He et al., 2021). Lastly, we adjusted for unobserved confounding bias in the linear mixed model, using a recently developed framework for causal inference based on null treatments approach (Miao et al., 2022). Designed to discern causal effects from multiple treatment variables within non-randomized, observational data, the null treatments approach hinges on the assumption that no fewer than half of the confounded treatments exert no causal influence on the outcome. It circumvents the need for prior knowledge regarding which treatments are null and eliminates the necessity for independence among treatments. Given our model’s inclusion of numerous treatment variables with shared variances due to the presence of unobserved confounders (Abdellaoui et al., 2022; Okbay et al., 2022)—including cognitive phenotypes PGS, family and neighborhood SES, positive family and school environments— we opted to employ this method.

Results

Demographics

Table 1 presents the demographic characteristics of the final samples. For multiethnic subjects (main analyses, n=6,602), 47.15% were female, and the parents of 70.21% were married. In European ancestry samples (sensitivity analyses, n=5,211), 46.71% were female, and the parents of 77.47% were married. Children of European ancestry showed significantly different marital status (p<0.0001), lower BMI (p<0.0001) and family history of psychiatric disorders (p<0.0001) compared to children of other genetic ancestries. Our linear mixed model and IGSCA analyses were adjusted using sex, age, marital status, BMI, family history of psychiatric disorders, and ABCD research sites as covariates.

Genetic influence on cognitive phenotypes correlates positively with intelligence and negatively with PLEs

As shown in Figure 2, higher PGSs of cognitive capacity phenotypes correlated significantly with higher intelligence (CP PGS: β=0.1113~0.1793; EA PGS: β=0.0699~0.1567). While CP PGS was associated only with lower baseline year Distress Score PLEs (β=-0.0323), EA PGS was associated with lower baseline year and follow-up PLEs of all measures (baseline: β=-0.0518~-0.0519; 1-year: β=-0.0423~-0.043; 2-year: β=-0.036~ −0.0463). No significant correlations were found between CP PGS and follow-up PLEs (p>0.05). The effects of EA PGS were larger on baseline year PLEs than follow-up PLEs. The effect sizes of EA PGS on children’s PLEs were larger than those of CP PGS (EA PGS: β=-0.036~-0.0519; CP PGS: β=-0.0323). (Table S1).

Linear models testing genetic, socioeconomic, and environmental factors associated with cognitive intelligence and PLEs.

Standardized coefficients of a linear mixed model with CP PGS (A, B) and EA PGS (C, D). The analyses included 6,602 samples of multiethnicity. Cognitive intelligence and PLEs correlated with the PGSs, residential disadvantage, positive environment, and family SES in the opposite directions. Error bars indicate 95% bootstrapped confidence intervals with 5,000 iterations. CP and EA denote cognitive performance and education attainment, respectively; PGS, polygenic scores; SES, socioeconomic status; PLEs, psychotic-like experiences; ADI, Area Deprivation Index; Poverty, percentage of individuals below −125% of the poverty level; Years, years of residence. (*p-FDR<0.05)

Family and neighborhood SES correlates positively with intelligence and negatively with PLEs

Parental education associated positively with all types of intelligence (β=0.0699~0.1745) and negatively with baseline year Total and Distress Score PLEs (β=-0.0528~-0.043), 1-year follow-up PLEs (β=-0.0538~-0.0449), and 2-year follow-up PLEs (β=-0.0459~-0.0389). Family income correlated positively with intelligence (β=0.0723~0.1365) and negatively with 2-year follow-up parent-rated PLEs (β=-0.0503~ - 0.0502). Family’s financial disadvantage correlated negatively with baseline year parent-rated PLEs (β=0.0726~0.0728), 1-year follow-up PLEs of all types (β=0.0307~0.0577), and 2-year follow-up PLEs of all types (β=0.0461~0.0581).

The ADI correlated significantly negatively with all types of intelligence (β=-0.0684~-0.054). Additionally, a higher ADI correlated significantly with higher baseline year PLEs (β=0.0587~0.0914), 1-year follow-up PLEs (β=0.0523~0.0613), and 2-year follow-up PLEs (β=0.0397~0.0449).

We found no significant associations of poverty with any of the target variables (p> 0.05). Years of residence correlated significantly with crystallized intelligence (β=0.035~0.0372) and baseline year Total Score PLEs (β=-0.029~-0.0273). (Table S1).

Positive family and school environments correlate positively with intelligence and negatively with the influence of PLEs

Positive parenting behaviors showed significant negative correlations with baseline year PLEs (β=-0.0702~-0.0419), 1-year follow-up PLEs (β=-0.0588~-0.0397), and 2-year follow-up PLEs (β=-0.0623~-0.0356) (Figure 2). Positive school environment was associated positively with total intelligence (β=0.0353~0.0397) and fluid intelligence (β=0.0514~0.0545) and negatively with all three measures of baseline year PLEs (β=-0.1193~-0.0468), 1-year follow-up PLEs (β=-0.1078~-0.04), and 2-year follow-up PLEs (β=-0.1068~-0.0586) (Table S1).

Structural Equation Modeling-Integrated Generalized Structured Component Analysis

The IGSCA model showed that intelligence mediated the effects of genes and environments on the risk for psychosis (PLEs) (Figure 3 and Table 2). Estimated factor loadings of latent factor variable and weights of component variables are presented in Table S2. The model showed a good model fit with a GFI of 0.9735, SRMR of 0.0359, and FIT value of 0.4912 (Cho et al., 2020). Intelligence was under significant direct influences of the cognitive phenotypes PGS (β=0.2427), family SES (β=0.2932), neighborhood SES (β=-0.1121), and positive environment (β=0.0268). Family SES and positive environment had significant negative direct effects on PLEs of all years. Cognitive phenotypes PGS and neighborhood SES showed no significant direct effects on any of the PLEs (p>0.05). Intelligence significantly mediated the effects of the PGS, family and neighborhood SES, and positive environment on PLEs of all years: the PGS (baseline year: β=-0.035; 1-year: β=-0.0355; 2-year: β=-0.0274), family SES (baseline year: β=-0.0423; 1-year: β=--0.0429; 2-year: β=-0.0331), neighborhood SES (baseline year: β=0.0162; 1-year: β=0.0164; 2-year: β=0.0126), and positive environment (baseline year: β=-0.0039; 1-year: β=-0.0039; 2-year: β=-0.003).

Direct/Indirect effects of gene-environment factors to cognitive and PLEs outcomes

A. Direct pathways from PGS, high family SES, low neighborhood SES, and positive environment to cognitive intelligence and PLEs. Standardized path coefficients are indicated on each path as direct effect estimates (significance level *p<0.05). B. Indirect pathways to PLEs via intelligence were significant for polygenic scores, high family SES, low neighborhood SES, and positive environment, indicating the significant mediating role of intelligence. C. Relative effect sizes of direct and indirect pathways within the total effects on PLEs. The standardized effect sizes of direct pathways are colored within each bar. In A-B, child sex, genetic ancestry, BMI, marital status, family history of psychiatric disorders, and ABCD research sites were included as covariates. CP PGS and EA PGS denote polygenic scores of cognitive performance and education attainment, respectively; SES, socioeconomic status; PLEs, psychotic-like experiences; Crystallized and Fluid, crystallized and fluid intelligence; ADI, Area Deprivation Index; Poverty, percentage of individuals below −125% of the poverty level; Years, years of residence.

Note: * indicates a statistically significant parameter estimate at α=0.05.

For all observed years, positive environment had largest total effects on PLEs (baseline year: β=-0.152; 1-year: β=-0.1316; 2-year: β=-0.1364), followed by family SES (baseline year: β=-0.1216; 1-year: β=-0.1119; 2-year: β=-0.1164), neighborhood SES (baseline year: β=0.0504; 1-year: β=0.0329; 2-year: β=0.0192), and PGS (baseline year: β=-0.0498; 1-year: β=-0.036; 2-year: β=-0.0365). The total effects of each indicator on PLEs were significant except for those of neighborhood SES (p>0.05).

Sensitivity Analyses

As sensitivity analyses, we reran our main analyses with adjustment for ethnic confounding, schizophrenia PGS, and unobserved confounding, respectively. Results of linear mixed model and IGSCA analyses were consistent (Table S3-S7). See Appendix S2 for detailed results.

Discussion

This study investigated the relationships of the genetic and environmental influences on the development of intelligence and the PLEs in youth, leveraging genetic data from the large epidemiological samples and a multi-level environmental (socioeconomic) data. Our results support the model that genetic factors (PGS for cognitive phenotypes), socioeconomic conditions, and family and school environments may influence cognitive intelligence in children, and this impact may lead to the individual variability of the current and future PLEs in children. Our analysis with integrated data shows the contributions of genetic and environmental factors, respectively, to cognitive and mental wellness in children, and further provides policy implications to improve them.

Our SEM shows that cognitive intelligence mediates the environmental and genetic influence on the current and future PLEs. The environmental factors (family SES, neighborhood SES, and positive parenting and schooling) and PGS of cognitive phenotypes exhibit significant indirect effects via cognitive intelligence on PLEs. Prior research identifying the mediation of cognitive intelligence focused on either genetic (Karcher, Paul, et al., 2021) or environmental factors (Lewis et al., 2020) alone. Studies with older clinical samples have shown that cognitive deficit may be a precursor for the onset of psychotic disorders (Eastvold et al., 2007; Fett et al., 2020; Vorstman et al., 2015). Our study advances this by demonstrating the integrated effects of genetic and environmental factors on PLEs through the cognitive intelligence in 9-11 years old children. Such comprehensive analysis contributes to assessing the relative importance of various factors influencing children’s cognition and mental health, and it can aid future studies designed for identifying health policy implications. Considering the directions and magnitudes of the effects, though the effects of PGS remain significant, aggregated effects of environmental factors account for much greater degrees on PLEs.

Our results of cognitive intelligence mediating the genetic and environmental effects on PLEs may be related to several potential mechanisms. Children raised in higher family SES may have sufficient nutrition and cognitive stimulants, whereas children living in deprived neighborhoods may be exposed to higher rates of crime, air pollution, and substance abuse (Lewis et al., 2020; Marshall et al., 2020; Taylor et al., 2020; Tomasi & Volkow, 2021). Environmental enrichment may be associated with longer periods of neural plasticity (e.g. myelination, maturation of brain circuitry), leading to higher cognitive ability and lower risk of mental disorders like PLEs (Tooley et al., 2021). This may be further linked to the cognitive reserve theory. The theory suggests that genetic influence for cognitive phenotypes and environmental enrichment promotes more efficient, flexible brain networks, which may lead to greater resilience against psychopathology (Stern, 2009). Indeed, prior clinical studies show the linkage between cognitive reserve and psychosis (Amoretti et al., 2018; Leeson et al., 2011).

Our results indicate that genetic influences on cognitive phenotypes are significantly linked to PLEs. PGSs for CP and EA were strongly correlated with PLEs (baseline year, 1-year follow-up, and 2-year follow-up). These associations were robust after adjustment for schizophrenia PGS, ethnic confounding and unobserved confounders. Cognitive phenotypes PGS generally show higher predictive performance than PGS of any other traits (Lee et al., 2018; Okbay et al., 2022; Plomin & von Stumm, 2018). Genetic variants associated with CP and EA are related to complex traits across the life span, including neuroticism, depressive symptoms, smoking in adulthood, cognitive decline at a later age (Joo et al., 2022), risk for Alzheimer’s disease (Lee et al., 2018; Okbay et al., 2022), brain volume, area, and thickness, as well as psychotic disorders (Karcher, Paul, et al., 2021). Prior gene expression studies suggest that polygenic signals for schizophrenia, bipolar disorder, and EA are significantly enriched in the central nervous system, particularly the cerebellum (Finucane et al., 2018). Our findings emphasize the importance of cognitive phenotypes PGS as a biomarker which not only implicates cognitive traits but also exhibits genetic overlap with the PLEs.

The differing magnitudes of the PGS impact between EA and CP warrant attention. The effects of the EA PGS on the PLEs of all years were 160.68%~371.67% larger than those of CP PGS. This discrepancy may result from that the larger sample size of EA GWAS than that of CP GWAS. Alternatively, the discrepancies in effect sizes may suggest different genetic compositions between EA and CP. Recent literature documents that more than half of the polygenic signal for EA is related to noncognitive and social skills required for successful educational attainment (Demange et al., 2021), whereas CP may rather be linked to cognitive skills. This observation also supports the well-established relationships of the EA PGS with socioeconomic and life-course outcomes (e.g., social mobility (Belsky et al., 2018), voter turnout (Aarøe et al., 2020), BMI, income, time spent watching television, geographic residence (Abdellaoui et al., 2022), and wealth inequality (Barth et al., 2020), which may be influenced by unobserved environmental factors (Young et al., 2019). In our analysis, the utilization of two PGSs a more comprehensive evaluation, contributing to an estimation of the genetic and environmental factors that attempted to minimize confounding bias.

Furthermore, the significant effects of cognitive phenotypes PGS on cognitive intelligence (β=0.0699~0.1793) remained robust and similar in magnitude after adjusting for genetic ancestry (β=0.0754~0.1866) and other (unobserved) confounding (β=0.0546~0.1776). As we controlled for family-, neighborhood-, and school-level environmental factors and unobserved confounders, our results may be interpreted as significant genetic influences on individual’s cognitive intelligence. This interpretation is supported by a recent study (Isungset et al., 2022): Despite of the socioeconomic differences in Norway (a typical social democratic welfare state) and the US (a typical liberal welfare state), the magnitudes of genetic influence on cognitive intelligence were similar (Norway: β=0.18; US: β=0.17). Cognitive phenotypes PGS is an important genetic factor across the nations and societies. Therefore, analyses omitting the genetic influence may be subject to overestimation of the socioeconomic impact (Plomin & von Stumm, 2018; Sariaslan et al., 2016).

This study shows that a high SES and positive environment, particularly positive parenting behavior and school environment, is associated with higher intelligence and a lower risk for PLEs in children. While prior research has emphasized the dominant role of family SES (e.g. family income) (Tomasi & Volkow, 2021), our SEM analyses (IGSCA) showed that positive environmental factors such as supportive parenting and schooling have a greater impact on children’s PLEs. Specifically, the effect sizes were the highest in supportive family and school environment, followed by family and neighborhood SES. Even after adjusting for genetic ancestry and unobserved confounders, the strong associations of positive parenting and schooling with higher intelligence and fewer PLEs remained significant. These findings suggest that interventions that target positive family and school environments may be particularly effective. Recent research supports this notion, showing that interventions that promote supportive parenting and inclusive school environments can improve neurocognitive ability, academic performance, and decrease risk behaviors such as drinking and emotional eating (Brody et al., 2017; Brody et al., 2019; Holmes et al., 2018).

Moreover, our results showed that positive parenting and schooling in baseline year were associated not only with baseline year PLEs but also with PLEs 1-2 years later. This is in line with prior research showing that intervention focused on parenting behavior and school environment have long-lasting positive effects that extend into adulthood and even across generations (Cunha & Heckman, 2007; Hill et al., 2020).

While policy implications in observational studies like ours might be limited, our findings show the importance of comprehensive approaches considering the entire ecosystem of children’s lives—including residential, family, and school environment—for future research aimed to enhance children’s intelligence and mental health. When we combine the total effect sizes of neighborhood and family SES, as well as positive school environment and parenting behavior (∑ |β| = 0.2718~0.3242), they considerably surpass the total effect sizes of cognitive phenotypes PGSs (|β| = 0.0359~0.0502). It has been suggested that a holistic and quantitative approach that takes into account the comprehensive ecosystem of family, school, and residential environments may ensure policy effects and efficient use of resources (Cree et al., 2018; Garner & Yogman, 2021; Shonkoff, 2012). For example, the Health Impact in 5 Years Initiative of the US Centers for Disease Control and Prevention (CDC, 2018) includes 14 evidence-based interventions, such as providing school-based prevention programs, public transportation, home improvement loans, and earned income tax credits, to tackle the social determinants of public health. Our study strengthens the idea that an interdisciplinary science-driven, coordinated approach to intervening in the select environmental factors may allow practical improvements in child development, particularly in those who are at a disadvantage.

Our study has some limitations. First, due to data availability constraints in the ABCD study, we only utilized baseline observations for NIH Toolbox cognitive intelligence, and we could not test whether PLEs might be a mediator of intelligence. Second, the generalizability of our findings may be limited since most of the participants included in our analysis are from European ancestry. Although the ABCD Study aimed to achieve its representativeness by recruiting from an array of school systems located around each of the 21 research sites, chosen for their diversity in geography, demographics, and socioeconomic status, it is not fully representative of the US population (Compton et al., 2019). Third, the duration of the follow-up period utilized in this study is relatively short (1-year and 2-year follow-up), which may limit the interpretability of our findings for understanding cognitive and psychiatric development during later childhood. Future research could potentially benefit from employing longer follow-up periods, as more follow-up observations are being collected in the ABCD Study. Fourth, while we used a wide range of statistical methods to adjust for confounding bias from observed and unobserved variables (e.g., genetic ancestry), we did not account for other types of potential bias such as sample selection bias. Fifth, despite a number of causal inference methods used in this study, the ABCD Study is a non-randomized dataset. Given the observational nature of the ABCD Study, interpreting our results as actual causality requires more caution. Finally, we did not include all important environmental variables, such as air pollution (Marshall et al., 2020) and social capital (Krabbendam, 2005), which are not collected in the ABCD study.

In conclusion, our study provides potential pathways of genetic factors of cognitive phenotypes and environmental factors of family, school, and neighborhood to cognitive and mental wellness in children. Our findings underscore the importance of a comprehensive approach that considers both biological and socioeconomic features in promoting young children’s cognitive ability and mental health. Given the importance of child development, it requires joint efforts from multiple disciplines.

Supporting information